---
title: "2022 16S ReAnalysis Diversity"
output: html_document
date: "2026-07-27"
---

> **BLOCKED -- open item, not yet resolved.** This Rmd needs
> `asv_otu_rl_2022_16S.csv`, a 2022 16S taxonomy table, and
> `Ossining_Lake_Metadata_2022.xlsx`. None of these exist anywhere in this
> repository's `data/` folder (confirmed by directory listing). The
> `ossining_meta.csv` massaging fix applied to the 2023 Rmds does NOT
> resolve this -- that file's 2022 rows have the environmental metadata,
> but the ASV count table and taxonomy for 2022 are missing entirely.
> `eval = FALSE` is set below so this doesn't fail knitting the rest of the
> project; remove it once the 2022 ASV/taxonomy data is added and the
> paths in the data-import chunk are updated to point at `data/`.

```{r setup, include=FALSE}
knitr::opts_chunk$set(eval = FALSE)
```

```{r}
#Installing packages(if not already installed)

#Installing package: dada2
#if (!requireNamespace("BiocManager", quietly = TRUE))
  #install.packages("BiocManager")
#BiocManager::install(version = "3.22")
#BiocManager::install("dada2")
#Installing package: phyloseq
#if (!require("BiocManager", quietly = TRUE))
  #install.packages("BiocManager")
#BiocManager::install("phyloseq")
#Installing package:DECIPHER
#if (!require("BiocManager", quietly = TRUE))
 #   install.packages("BiocManager")
#BiocManager::install("DECIPHER")
#Installing package: microbiome
#if (!require("BiocManager", quietly = TRUE))
  #install.packages("BiocManager")
#BiocManager::install("microbiome")
#install.packages ("phangorn")
#install.packages("svglite")
#install.packages("patchwork")

# Install from GitHub
if (!requireNamespace("remotes", quietly = TRUE)){
  install.packages("remotes")
}
remotes::install_github("vmikk/metagMisc")

#installing dada2 and DECIPHER
options(BioC_mirror = "https://packagemanager.posit.co/bioconductor")
options(BIOCONDUCTOR_CONFIG_FILE = "https://packagemanager.posit.co/bioconductor/config.yaml")

if (!requireNamespace("BiocManager", quietly = TRUE))
  install.packages("BiocManager")

BiocManager::install(c("dada2", "DECIPHER"))
```

```{r}
#------- load packages -------#
#Load libraries
library("ggplot2")
library("vegan")
library("dplyr")
library("scales")
library("grid")
library("reshape2")
library("phyloseq")
library("phangorn")
library("DECIPHER")
library("dada2")
library("microbiome")
library("RColorBrewer")
library("stringr")
library("readxl")
library("svglite")
library("patchwork")
library("metagMisc")
```

```{r}
#------- import data -------#
# Sequence count table
otutab <- read.csv("~/Ossining Manuscript/Re-Analysis/2022 16S/asv_otu_rl_2022_16S.csv", check.names = F)
rownames(otutab)<- otutab[,1]
otutab <- otutab[,2:30]
otutab <- otu_table(otutab, taxa_are_rows = T)
# Sequence taxonomy classifications
tax <- read.csv("~/Ossining Manuscript/Re-Analysis/taxonomy_16S-2022.csv", header=T)
rownames(tax)<- taxa_names(otutab)
tax <- tax[,2:9]
tax <- as.matrix(tax)
tax <- tax_table(tax)
taxa_names(tax) <- taxa_names(otutab)
# Sample metadata
meta <- read_xlsx("~/Ossining/Data Analysis/Ossining_Lake_Metadata_2022.xlsx")
meta <- sample_data(meta)
rownames(meta) <- meta$Name
# Load table, tax, and metadata into phyloseq object
physeq_16S22 <- merge_phyloseq(otutab, tax, meta)

# remove samples with zero counts
physeq_16S22 <- prune_samples(sample_sums(physeq_16S22) >= 1, physeq_16S22)

# remove chloroplasts
physeq_16S22 <- subset_taxa(physeq_16S22, Order!="Chloroplast")
```

```{r}
#------- alpha diversity measures -------#
library("microbiome")
library("knitr")
alpha16S22 <- microbiome::alpha(physeq_16S22)
write.csv(alpha16S22, "~/Ossining Manuscript/Re-Analysis/Diversity R/ossining22-16S_alpha-diversity-2.csv")

#------- plot richness -------#
jpeg("richplot16S_22-2.jpg",width = 10, height =8, units = "in", res = 300)
plot_richness(physeq_16S22, x = "Date_Sam", measures="Observed") +
  geom_boxplot() +
  theme_bw() + theme(axis.text.x=element_text(angle=60, hjust=1)) + 
  theme(axis.text = element_text(size = 14))+
  theme(axis.title.x=element_blank())
dev.off()

jpeg("richplot16S_22_lake-2.jpg",width = 10, height =8, units = "in", res = 300)
plot_richness(physeq_16S22, x = "Date_Sam", measures="Observed", title = "16S 2022 richness") +
  geom_boxplot() +
  theme_bw() + theme(axis.text.x=element_text(angle=60, hjust=1)) + 
  theme(axis.text = element_text(size = 14))+
  theme(axis.title.x=element_blank()) +facet_grid(.~Lake, scales="free")
dev.off()

rich.plot_16S22 <- plot_richness(physeq_16S22, x = "Date_Sam", measures="Observed") +
  geom_boxplot() +
  theme_bw() + theme(axis.text.x=element_text(angle=60, hjust=1)) + 
  theme(axis.text = element_text(size = 14))+
  theme(axis.title.x=element_blank())

#------- Phylum bar plot -------#
p_16S22 <- subset_taxa(physeq_16S22) %>%
  tax_glom(taxrank = "Phylum") %>%                     #agglomerate to phylum level
  transform_sample_counts(function(x){x/sum(x)}) %>%   #transform to relative abundance
  psmelt()
write.csv(p_16S22, "~/Ossining Manuscript/Re-Analysis/Diversity R/ossining22-16S_phylum-2.csv")

#library(dplyr)
relabund_table_16S2022Phylum <- p_16S22 %>%
  group_by(Lake, Month, Date_Sam, Phylum) %>%
  summarize(RelAbund = sum(Abundance), .groups = "drop")

write.csv(relabund_table_16S2022Phylum,
          "~/Ossining Manuscript/Re-Analysis/Diversity R/ossining22-16S_relative_abundance_phylum-2.csv",
          row.names = FALSE)

#------- Build a master color palette based on 2023 data -------#
all_phyla_2023 <- sort(unique(p_16S23$Phylum))
all_phyla_2022 <- sort(unique(p_16S22$Phylum))

extra_2022_only <- setdiff(all_phyla_2022, all_phyla_2023)

# Combine so 2023 phyla always get the same colors, extras get appended
all_phyla_combined <- c(all_phyla_2023, extra_2022_only)

getPalette <- colorRampPalette(brewer.pal(12, "Paired"))
phylum_colors <- setNames(getPalette(length(all_phyla_combined)), all_phyla_combined)
#-------------------------------------------------------------------------------#

phyplot_16S22 <- ggplot(arrange(p_16S22, Phylum), aes(x=Date_Sam, y=Abundance, fill=Phylum)) +
  geom_bar(stat = "identity") + 
  theme_bw() + 
  theme(axis.title.x = element_blank(),
        plot.title = element_text(size=10, hjust = 0.5), 
        legend.text=element_text(size=10)) +
  ylab("Relative Read Abundance") +
  theme(axis.text = element_text(size = 10))+
  scale_fill_manual(values = phylum_colors) +
  scale_y_continuous(limits = c(0,1), expand = c(0,0))+
  guides(fill=guide_legend(ncol = 2))+
  theme(axis.text.x = element_text(angle = 45, hjust=1))

phyplot_16S22 +facet_grid(.~Lake, scales = "free_x", space = "free_x")+
  theme(strip.text.x = element_text(size = 9))

#SAVE PLOTS:
jpeg("phyplot16S_22-2.jpg",width = 10, height =8, units = "in", res = 300)
ggplot(arrange(p_16S22, Phylum), aes(x=Date_Sam, y=Abundance, fill=Phylum)) +
  geom_bar(stat = "identity") +
  theme_bw() +
  theme(axis.title.x = element_blank(),
        plot.title = element_text(size=10, hjust = 0.5), 
        legend.text=element_text(size=10)) +
  ylab("Relative Read Abundance") +
  theme(axis.text = element_text(size = 10))+
  scale_fill_manual(values = phylum_colors) +
  scale_y_continuous(limits = c(0,1), expand = c(0,0))+
  guides(fill=guide_legend(ncol = 2))+
  theme(axis.text.x = element_text(angle = 45, hjust=1))
dev.off()

jpeg("lake_phyplot16S_22-2.jpg",width = 10, height =8, units = "in", res = 300)
phyplot_16S22 +facet_grid(.~Lake, scales = "free_x", space = "free_x")+
  theme(strip.text.x = element_text(size = 9)) + plot_annotation ("16S 2022 Relative Read Abundance")
dev.off()

#save as svg
ggsave("lake_phyplot16S_22-2.svg",width = 10, height =8, units = "in")
phyplot_16S22 +facet_grid(.~Lake, scales="free_x", space = "free_x")+
  theme(strip.text.x = element_text(size = 9)) + plot_annotation ("16S 2022 Relative Read Abundance")
dev.off()
```

```{r}

#------- Cyanobacteria bar plot -------#
cyano_16S22 <- subset_taxa(physeq_16S22, Phylum=="Cyanobacteriota") %>%
  tax_glom(taxrank = "Taxa") %>%                     #agglomerate to edited genus level(Taxa)
  transform_sample_counts(function(x){x/sum(x)}) %>%   #transform to relative abundance
  psmelt()%>%                                         # melt to long format
  filter(Abundance > 0.01) 
write.csv(cyano_16S22, "~/Ossining Manuscript/Re-Analysis/Diversity R/ossining22-16S_cyanobacteria-2.csv")

#library(dplyr)
relabund_table_16S2022Cyano <- cyano_16S22 %>%
  group_by(Lake, Month, Date_Sam, Taxa) %>%
  summarize(RelAbund = sum(Abundance), .groups = "drop")

write.csv(relabund_table_16S2022Cyano,
          "~/Ossining Manuscript/Re-Analysis/Diversity R/ossining22-16S_relative_abundance_cyano-2.csv",
          row.names = FALSE)

#------- Build a master color palette based on 2023 data -------#
all_cyano_2023 <- sort(unique(cyano_16S23$Taxa))
all_cyano_2022 <- sort(unique(cyano_16S22$Taxa))

extra_cyano_2022_only <- setdiff(all_cyano_2022, all_cyano_2023)

# Combine so 2023 phyla always get the same colors, extras get appended
all_cyano_combined <- c(all_cyano_2023,extra_cyano_2022_only)

getPalette <- colorRampPalette(brewer.pal(8, "Set1"))
cyano_colors <- setNames(getPalette(length(all_cyano_combined)), all_cyano_combined)
#-------------------------------------------------------------------------------#

cyano_plot_16S22 <- ggplot(arrange(cyano_16S22, Taxa), aes(x=Date_Sam, y=Abundance, fill=Taxa)) +
  geom_bar(stat = "identity") + theme_bw() +
  theme(plot.title = element_text(size=10, hjust = 0.5), 
        axis.title.x = element_blank(),
        legend.text=element_text(size=10),
        legend.position = "right") +
  theme(axis.text = element_text(size = 10))+
  ylab("Relative Read Abundance") +
  scale_fill_manual(values = cyano_colors) +
  scale_y_continuous(limits = c(0,1.001), expand = c(0,0))+
  guides(fill=guide_legend(ncol = 1))+
  theme(axis.text.x = element_text(angle = 45, hjust=1))

cyano_plot_16S22 +facet_grid(.~Lake, scales = "free_x", space = "free_x")+
  theme(strip.text.x = element_text(size = 9))

jpeg("cyanoplot16S_22-2.jpg",width = 10, height =8, units = "in", res = 300)
ggplot(arrange(cyano_16S22, Taxa), aes(x=Date_Sam, y=Abundance, fill=Taxa)) +
  geom_bar(stat = "identity") + theme_bw() +
  theme(plot.title = element_text(size=10, hjust = 0.5), 
        axis.title.x = element_blank(),
        legend.text=element_text(size=10),
        legend.position = "right") +
  theme(axis.text = element_text(size = 10))+
  ylab("Relative Read Abundance") +
  scale_fill_manual(values = cyano_colors) +
  scale_y_continuous(limits = c(0,1.001), expand = c(0,0))+
  guides(fill=guide_legend(ncol = 1))+
  theme(axis.text.x = element_text(angle = 45, hjust=1))
dev.off()

jpeg("lake_cyanoplot16S_22-2.jpg",width = 10, height =8, units = "in", res = 300)
cyano_plot_16S22 +facet_grid(.~Lake, scales = "free_x", space = "free_x")+
  theme(strip.text.x = element_text(size = 9)) + plot_annotation ("16S 2022 Cyanobacteria Relative Read Abundance")
dev.off()

#save as svg to edit colors
ggsave("lake_cyanoplot16S_22-2.svg",width = 10, height =8, units = "in")
cyano_plot_16S22 +facet_grid(.~Lake, scales = "free_x", space = "free_x")+
  theme(strip.text.x = element_text(size = 9)) + plot_annotation ("16S 2022 Cyanobacteria Relative Read Abundance")
dev.off()
```

```{r}

#------- PCoA  -------#
#subset by lake
LM_16S22 <- subset_samples(physeq_16S22, Lake =="Lake Mohegan")
LM_16S22
#otu_table()   OTU Table:         [ 6467 taxa and 11 samples ]
#sample_data() Sample Data:       [ 11 samples by 12 sample variables ]
#tax_table()   Taxonomy Table:    [ 6467 taxa by 8 taxonomic ranks ]

#scale to even depth
LM_16S22rare <- rarefy_even_depth(LM_16S22)
LM_16S22rare
#otu_table()   OTU Table:         [ 1894 taxa and 11 samples ]
#sample_data() Sample Data:       [ 11 samples by 12 sample variables ]
#tax_table()   Taxonomy Table:    [ 1894 taxa by 8 taxonomic ranks ]
sample_sums(LM_16S22rare) #4851

#PCoA of Mohegan Lake
pcoa_LM_16S22 <- ordinate(physeq=LM_16S22rare, method = "PCoA", distance = "bray")

jpeg("LM_pcoa_plot16S_22-2.jpg",width = 10, height =8, units = "in", res = 300)
plot_scree(pcoa_LM_16S22) 
dev.off()

ord_LM_16S22 <- plot_ordination(physeq=LM_16S22rare, ordination = pcoa_LM_16S22, color = "Month", 
                title = "Lake Mohegan") +
  geom_point(aes(color = Month), alpha = 0.75, size = 4)+
  theme_bw()+
  theme(text = element_text(size = 14))+
  guides(color = guide_legend(title = "Month", ncol = 1))

##rippowam lake
#subset by lake
LR_16S22 <- subset_samples(physeq_16S22, Lake =="Lake Rippowam")
LR_16S22
#otu_table()   OTU Table:         [ 6467 taxa and 6 samples ]
#sample_data() Sample Data:       [ 6 samples by 12 sample variables ]
#tax_table()   Taxonomy Table:    [ 6467 taxa by 8 taxonomic ranks ]

#scale to even depth
LR_16S22rare <- rarefy_even_depth(LR_16S22)
LR_16S22rare
#otu_table()   OTU Table:         [ 799 taxa and 6 samples ]
#sample_data() Sample Data:       [ 6 samples by 12 sample variables ]
#tax_table()   Taxonomy Table:    [ 799 taxa by 8 taxonomic ranks ]
sample_sums(LR_16S22rare) #2754

#PCoA of Rippowan Lake
pcoa_LR_16S22 <- ordinate(physeq=LR_16S22rare, method = "PCoA", distance = "bray")

jpeg("LR_pcoa_plot16S_22-2.jpg",width = 10, height =8, units = "in", res = 300)
plot_scree(pcoa_LR_16S22) 
dev.off()

ord_LR_16S22 <- plot_ordination(physeq=LR_16S22rare, ordination = pcoa_LR_16S22, color = "Month", 
                title = "Lake Rippowam") +
  geom_point(aes(color = Month), alpha = 0.75, size = 4)+
  theme_bw()+
  theme(text = element_text(size = 14))+
  guides(color = guide_legend(title = "Month", ncol = 1))

###Teatown Lake
#subset by lake
TTL_16S22 <- subset_samples(physeq_16S22, Lake =="Teatown Lake")
TTL_16S22
#otu_table()   OTU Table:         [ 6467 taxa and 12 samples ]
#sample_data() Sample Data:       [ 12 samples by 12 sample variables ]
#tax_table()   Taxonomy Table:    [ 6467 taxa by 8 taxonomic ranks ]

#scale to even depth
TTL_16S22rare <- rarefy_even_depth(TTL_16S22)
TTL_16S22rare
#otu_table()   OTU Table:         [ 3292 taxa and 12 samples ]
#sample_data() Sample Data:       [ 12 samples by 12 sample variables ]
#tax_table()   Taxonomy Table:    [ 3292 taxa by 8 taxonomic ranks ]
sample_sums(TTL_16S22rare) #30196

#PCoA of Teatown Lake
pcoa_TTL_16S22 <- ordinate(physeq=TTL_16S22rare, method = "PCoA", distance = "bray")

jpeg("TTL_pcoa_plot16S_22-2.jpg",width = 10, height =8, units = "in", res = 300)
plot_scree(pcoa_LR_16S22) 
dev.off()

ord_TTL_16S22 <- plot_ordination(physeq=TTL_16S22rare, ordination = pcoa_TTL_16S22, color = "Month", title = "Teatown Lake") +
  geom_point(aes(color = Month), alpha = 0.75, size = 4)+
  theme_bw()+
  theme(text = element_text(size = 14))+
  guides(color = guide_legend(title = "Month", ncol = 1))
#PATCHWORK FOR LAKE
jpeg("ordination_pcoa_plot16S_22-2.jpg",width = 10, height =8, units = "in", res = 300)
ord_LM_16S22 / ord_LR_16S22 / ord_TTL_16S22 + plot_annotation ("PCoA of 16S 2022 Data")
dev.off()

ggsave("ordination_pcoa_plot16S_22-2.svg",width = 10, height =8, units = "in")
ord_LM_16S22 / ord_LR_16S22 / ord_TTL_16S22 + plot_annotation ("PCoA of 16S 2022 Data")
dev.off()

#save jpeg each individual lake if needed EX:
##jpeg("LM_ordination_pcoa_plot16S_22.jpg",width = 10, height =8, units = "in", res = 300)
##plot_ordination(physeq=LM_16S22rare, ordination = pcoa_LM_16S22, color = "Month",  title = "PCoA of LM 16S Data 2022") + geom_point(aes(color = Month), alpha = 0.75, size = 4)+ theme_bw()+ theme(text = element_text(size = 14))+ guides(color = guide_legend(title = "Month", ncol = 1))

```

```{r}

#------- PCoA cyanobacteria only  -------#
# subset just cyanobacteria
cb_16S22 <- subset_taxa(physeq_16S22, Phylum=="Cyanobacteriota")
alphacyano22 <- microbiome::alpha(cb_16S22)
write.csv(alphacyano22, "~/Ossining Manuscript/Re-Analysis/Diversity R/ossining22-16S-cyano_alpha-diversity-2.csv")

# subset by Lake (Mohegan = LM)
LMcb_16S22 <- subset_samples(cb_16S22, Lake =="Lake Mohegan")
LMcb_16S22
#otu_table()   OTU Table:         [ 1848 taxa and 11 samples ]
#sample_data() Sample Data:       [ 11 samples by 12 sample variables ]
#tax_table()   Taxonomy Table:    [ 1848 taxa by 8 taxonomic ranks ]

#scale to even depth
LMcbrare_16S22 <- rarefy_even_depth(LMcb_16S22)
LMcbrare_16S22
#otu_table()   OTU Table:         [ 710 taxa and 11 samples ]
#sample_data() Sample Data:       [ 11 samples by 12 sample variables ]
#tax_table()   Taxonomy Table:    [ 710 taxa by 8 taxonomic ranks ]
sample_sums(LMcbrare_16S22) #3277

#PCoA cyano only
LMcbpca_16S22 <- ordinate(LMcbrare_16S22, "PCoA")

jpeg("LM_cbca_plot16S_22-2.jpg",width = 10, height =8, units = "in", res = 300)
plot_scree(LMcbpca_16S22)
dev.off()

#ordination plot
cb_ord_LM_16S22 <- plot_ordination(LMcbrare_16S22, LMcbpca_16S22, color = "Month",
                title = "Lake Mohegan")+
  geom_point(aes(color = Month), alpha = 0.75, size = 4)+
  theme_bw()+
  theme(text = element_text(size = 14))+
  guides(color = guide_legend(title = "Month", ncol = 1))

##Rippowam Lake
# subset by Lake (Rippowan = LR)
LRcb_16S22 <- subset_samples(cb_16S22, Lake =="Lake Rippowam")
LRcb_16S22
#otu_table()   OTU Table:         [ 1848 taxa and 6 samples ]
#sample_data() Sample Data:       [ 6 samples by 12 sample variables ]
#tax_table()   Taxonomy Table:    [ 1848 taxa by 8 taxonomic ranks ]

#scale to even depth
LRcbrare_16S22 <- rarefy_even_depth(LRcb_16S22)
LRcbrare_16S22
#otu_table()   OTU Table:         [ 146 taxa and 6 samples ]
#sample_data() Sample Data:       [ 6 samples by 12 sample variables ]
#tax_table()   Taxonomy Table:    [ 146 taxa by 8 taxonomic ranks ]
sample_sums(LRcbrare_16S22)  #1230

#PCoA cyano only
LRcbpca_16S22 <- ordinate(LRcbrare_16S22, "PCoA")

jpeg("LR_cbca_plot16S_22-2.jpg",width = 10, height =8, units = "in", res = 300)
plot_scree(LRcbpca_16S22)
dev.off()

#ordination plot
cb_ord_LR_16S22 <- plot_ordination(LRcbrare_16S22, LRcbpca_16S22, color = "Month",
                title = "Lake Rippowam")+
  geom_point(aes(color = Month), alpha = 0.75, size = 4)+
  theme_bw()+
  theme(text = element_text(size = 14))+
  guides(color = guide_legend(title = "Month", ncol = 1))

###Teatown Lake
# subset by Lake (Teatown = TTL)
TTLcb_16S22 <- subset_samples(cb_16S22, Lake =="Teatown Lake")
TTLcb_16S22
#otu_table()   OTU Table:         [ 1848 taxa and 12 samples ]
#sample_data() Sample Data:       [ 12 samples by 12 sample variables ]
#tax_table()   Taxonomy Table:    [ 1848 taxa by 8 taxonomic ranks ]

#scale to even depth
TTLcbrare_16S22 <- rarefy_even_depth(TTLcb_16S22)
TTLcbrare_16S22
#otu_table()   OTU Table:         [ 1009 taxa and 12 samples ]
#sample_data() Sample Data:       [ 12 samples by 12 sample variables ]
#tax_table()   Taxonomy Table:    [ 1009 taxa by 8 taxonomic ranks ]
sample_sums(TTLcbrare_16S22)  #13725

#PCoA cyano only
TTLcbpca_16S22 <- ordinate(TTLcbrare_16S22, "PCoA")

jpeg("TTL_cbca_plot16S_22-2.jpg",width = 10, height =8, units = "in", res = 300)
plot_scree(TTLcbpca_16S22)
dev.off()

#ordination plot
cb_ord_TTL_16S22 <- plot_ordination(TTLcbrare_16S22, TTLcbpca_16S22, color = "Month",
                title = "Teatown Lake")+
  geom_point(aes(color = Month), alpha = 0.75, size = 4)+
  theme_bw()+
  theme(text = element_text(size = 14))+
  guides(color = guide_legend(title = "Month", ncol = 1))

#PATCHWORK TO GROUP LAKES
jpeg("cyano_ordination_pcoa_plot16S_22-2.jpg",width = 10, height =8, units = "in", res = 300)
cb_ord_LM_16S22 / cb_ord_LR_16S22 / cb_ord_TTL_16S22 + plot_annotation ("Cyanobacteria PCoA of 16S 2022 Data")
dev.off()

ggsave("cyano_ordination_pcoa_plot16S_22-2.svg",width = 10, height =8, units = "in")
cb_ord_LM_16S22 / cb_ord_LR_16S22 / cb_ord_TTL_16S22 + plot_annotation ("Cyanobacteria PCoA of 16S 2022 Data")
dev.off()
```

```{r}
#------PERMANOVA--------#
#FOR MOHEGAN
#bray curtis on rarified physeq
LMbray16S22 <- phyloseq::distance(LM_16S22rare, method="bray")
#make physeq a data frame
LMdf16S22 <- data.frame(sample_data(LM_16S22rare))
#PERMANOVA with bray curtis in formula and data as dataframe
LMperm16S22 <- adonis2(LMbray16S22~Month, data=LMdf16S22)
#check output
LMperm16S22
#significant!

#FOR RIPPOWAM
#bray curtis on rarified physeq
LRbray16S22 <- phyloseq::distance(LR_16S22rare, method="bray")
#make physeq a data frame
LRdf16S22 <- data.frame(sample_data(LR_16S22rare))
#PERMANOVA with bray curtis in formula and data as dataframe
LRperm16S22 <- adonis2(LRbray16S22~Month, data=LRdf16S22)
#check output
LRperm16S22
#NOT significant

#FOR TEATOWN
#bray curtis on rarified physeq
TTLbray16S22 <- phyloseq::distance(TTL_16S22rare, method="bray")
#make physeq a data frame
TTLdf16S22 <- data.frame(sample_data(TTL_16S22rare))
#PERMANOVA with bray curtis in formula and data as dataframe
TTLperm16S22 <- adonis2(TTLbray16S22~Month, data=TTLdf16S22)
#check output
TTLperm16S22
#NOT significant

#-----Cyano PERMANOVA-----
#FOR MOHEGAN
#bray curtis on rarified physeq
cLMbray16S22 <- phyloseq::distance(LMcbrare_16S22, method="bray")
#make physeq a data frame
cLMdf16S22 <- data.frame(sample_data(LMcbrare_16S22))
#PERMANOVA with bray curtis in formula and data as dataframe
cLMperm16S22 <- adonis2(cLMbray16S22~Month, data=cLMdf16S22)
#check output
cLMperm16S22
#significant!

#FOR RIPPOWAM
#bray curtis on rarified physeq
cLRbray16S22 <- phyloseq::distance(LRcbrare_16S22, method="bray")
#make physeq a data frame
cLRdf16S22 <- data.frame(sample_data(LRcbrare_16S22))
#PERMANOVA with bray curtis in formula and data as dataframe
cLRperm16S22 <- adonis2(cLRbray16S22~Month, data=cLRdf16S22)
#check output
cLRperm16S22
#NOT significant

#FOR TEATOWN
#bray curtis on rarified physeq
cTTLbray16S22 <- phyloseq::distance(TTLcbrare_16S22, method="bray")
#make physeq a data frame
cTTLdf16S22 <- data.frame(sample_data(TTLcbrare_16S22))
#PERMANOVA with bray curtis in formula and data as dataframe
cTTLperm16S22 <- adonis2(cTTLbray16S22~Month, data=cTTLdf16S22)
#check output
cTTLperm16S22
#NOT significant

#ALL LAKES: physeq_16S22
#scale to even depth
Lake22rare <- rarefy_even_depth(physeq_16S22)
Lake22rare
#otu_table()   OTU Table:         [ 4956 taxa and 29 samples ]
#sample_data() Sample Data:       [ 29 samples by 12 sample variables ]
#tax_table()   Taxonomy Table:    [ 4956 taxa by 8 taxonomic ranks ]
sample_sums(Lake22rare) #2754

#bray curtis on rarified physeq
Lake22bray <- phyloseq::distance(Lake22rare, method="bray")
#make physeq a data frame
Lake22df <- data.frame(sample_data(Lake22rare))

#PERMANOVA with bray curtis in formula and data as dataframe
Lake22perm <- adonis2(Lake22bray~Lake*Month, data=Lake22df, by="terms")
#check output
Lake22perm
#NOT significant month, significant Lake, Lake*month

#ALL LAKES CYANO cb_16S22
#scale to even depth
cb_16S22rare <- rarefy_even_depth(cb_16S22)
cb_16S22rare
#otu_table()   OTU Table:         [ 1568 taxa and 29 samples ]
#sample_data() Sample Data:       [ 29 samples by 12 sample variables ]
#tax_table()   Taxonomy Table:    [ 1568 taxa by 8 taxonomic ranks ]
sample_sums(cb_16S22rare) #1230

#bray curtis on rarified physeq
cLake22bray <- phyloseq::distance(cb_16S22rare, method="bray")
#make physeq a data frame
cLake22df <- data.frame(sample_data(cb_16S22rare))
#PERMANOVA with bray curtis in formula and data as dataframe
cLake22perm <- adonis2(cLake22bray~Lake*Month, data=cLake22df, by="terms")
#check output
cLake22perm
#NOT significant month, significant lake and lake*month

``` 


```{r}
#PAIRWISE TESTING FOLLOWING PERMANOVA

#LM significant -> pairwise comparison post hoc
LMdf16S22$Month=as.factor(LMdf16S22$Month)
LMpairwise16S22 <- adonis_pairwise(x=LMdf16S22, dd=LMbray16S22, group.var="Month")
LMpairwise16S22

#LM cyano significant -> pairwise comparison post hoc
cLMdf16S22$Month=as.factor(cLMdf16S22$Month)
cLMpairwise16S22 <- adonis_pairwise(x=cLMdf16S22, dd=cLMbray16S22, group.var="Month")
cLMpairwise16S22

#Interaction significant lake only -> pairwise comparison post hoc
Lake22df$Lake=as.factor(Lake22df$Lake)
Lake22pairwise <- adonis_pairwise(x=Lake22df, dd=Lake22bray, group.var="Lake")
Lake22pairwise
#LR/TTL significant

#cyano Interaction significant lake only -> pairwise comparison post hoc
cLake22df$Lake=as.factor(cLake22df$Lake)
cLake22pairwise <- adonis_pairwise(x=cLake22df, dd=cLake22bray, group.var="Lake")
cLake22pairwise
#LR/TTL significant
```

```{r}
#------HOMOGENEITY OF MULTIVARIATE DISPERSION (PERMDISP)------#
# Run before interpreting each PERMANOVA above

#===== 16S FULL COMMUNITY, 2022 =====#

#FOR MOHEGAN
LMdisp16S22 <- betadisper(LMbray16S22, LMdf16S22$Month)
permutest(LMdisp16S22, permutations = 999)
# not significant

#FOR RIPPOWAM
LRdisp16S22 <- betadisper(LRbray16S22, LRdf16S22$Month)
permutest(LRdisp16S22, permutations = 999)
# not significant

#FOR TEATOWN
TTLdisp16S22 <- betadisper(TTLbray16S22, TTLdf16S22$Month)
permutest(TTLdisp16S22, permutations = 999)
# SIGNIFICANT
TukeyHSD(TTLdisp16S22)
# June-aug, june-july, sept-june significant


#===== ALL LAKES: physeq_16S22 (Lake*Month) =====#

#dispersion by Lake (main effect) -- significant PERMANOVA term, check this
Lake22disp_Lake <- betadisper(Lake22bray, Lake22df$Lake)
permutest(Lake22disp_Lake, permutations = 999)
# not significant

#dispersion by Month (main effect) -- PERMANOVA NOT significant here, but still good practice to check
Lake22disp_Month <- betadisper(Lake22bray, Lake22df$Month)
permutest(Lake22disp_Month, permutations = 999)
# not significant

#dispersion by Lake:Month interaction (combined factor) -- significant PERMANOVA term, check this
Lake22df$LakeMonth <- interaction(Lake22df$Lake, Lake22df$Month)
Lake22disp_LakeMonth <- betadisper(Lake22bray, Lake22df$LakeMonth)
permutest(Lake22disp_LakeMonth, permutations = 999)
# SIGNIFICANT
TukeyHSD(Lake22disp_LakeMonth)$group[, "p adj"]
#none were significant

#===== CYANO 16S, 2022 =====#

#FOR MOHEGAN
cLMdisp16S22 <- betadisper(cLMbray16S22, cLMdf16S22$Month)
permutest(cLMdisp16S22, permutations = 999)
#SIGNIFICANT
TukeyHSD(cLMdisp16S22)
# June-aug and sept-june significant

#FOR RIPPOWAM
cLRdisp16S22 <- betadisper(cLRbray16S22, cLRdf16S22$Month)
permutest(cLRdisp16S22, permutations = 999)
# not significant

#FOR TEATOWN
cTTLdisp16S22 <- betadisper(cTTLbray16S22, cTTLdf16S22$Month)
permutest(cTTLdisp16S22, permutations = 999)
#SIGNIFICANT
TukeyHSD(cTTLdisp16S22)
# June-aug, june-july, and sept-june significant

#===== ALL LAKES CYANO: cb_16S22 (Lake*Month) =====#

#dispersion by Lake (main effect) -- significant PERMANOVA term, check this
cLake22disp_Lake <- betadisper(cLake22bray, cLake22df$Lake)
permutest(cLake22disp_Lake, permutations = 999)
# not significant

#dispersion by Month (main effect) -- PERMANOVA NOT significant here, but still good practice to check
cLake22disp_Month <- betadisper(cLake22bray, cLake22df$Month)
permutest(cLake22disp_Month, permutations = 999)
# not significant

#dispersion by Lake:Month interaction (combined factor) -- significant PERMANOVA term, check this
cLake22df$LakeMonth <- interaction(cLake22df$Lake, cLake22df$Month)
cLake22disp_LakeMonth <- betadisper(cLake22bray, cLake22df$LakeMonth)
permutest(cLake22disp_LakeMonth, permutations = 999)
# not significant

```
