---
title: "2023 16S DADA2 Processing"
output: html_document
---

> **Reference only -- not executable outside the originating HPC environment.**
> This Rmd documents the exact DADA2 workflow used to generate
> `data/asv_otu_rl_2023_16S.csv` / `data/asv_tax_rl_2023_16S.csv` from raw
> sequences. It reads/writes absolute HPC cluster paths
> (`/../../projects/gmgi/...`) and reference databases (CyanoSeq, SILVA)
> that are not part of this repository, and raw sequences are not included
> here (see README "Data availability" for the BioProject accession). All
> chunks are set to `eval = FALSE` below so that knitting this file
> documents the method without attempting to re-run it.

```{r setup, include=FALSE}
knitr::opts_chunk$set(eval = FALSE)
```

Installing packages (if not already installed)
```{r}
#Installing package: dada2
if (!requireNamespace("BiocManager", quietly = TRUE))
  install.packages("BiocManager")
BiocManager::install("dada2", version = "3.16")

#Installing package: phyloseq
if (!require("BiocManager", quietly = TRUE))
  install.packages("BiocManager")
BiocManager::install("phyloseq")

#Installing package: Biostrings
if (!require("BiocManager", quietly = TRUE))
  install.packages("BiocManager")
BiocManager::install("Biostrings")

```

Loading libraries and path
```{r}
# load required programs
library(dada2)
library(phyloseq)
library(Biostrings)
library(ggplot2)

# CHANGE to the directory containing the trimmed unzipped fastq files.
path_test <- "/../../projects/gmgi/ecosystem-diversity/ossininglakes/Re-Run_2026/2023/16S/cutadapt"

#check all files are in the path
list.files(path_test)

# Forward and reverse fastq file names have format: SAMPLENAME_R1_trimmed.fastq and SAMPLENAME_R2_trimmed.fastq
fnFs_test <- sort(list.files(path_test, pattern="_R1_trimmed.fastq.gz", full.names = TRUE))
fnRs_test <- sort(list.files(path_test, pattern="_R2_trimmed.fastq.gz", full.names = TRUE))

# Extract sample names, assuming file names have format: SAMPLENAME_XXX.fastq
sample.names.test <- sapply(strsplit(basename(fnFs_test), "_"), `[`, 1)

# visualize the quality profiles of the forward reads (note where to truncate)
jpeg("fnFsQuality16S2023_test.jpg",width = 20, height =17, units = "in", res = 300)
plotQualityProfile(fnFs_test[1:45])
#truncate at 230

# visualize the quality profiles of the reverse reads (note where to truncate)
jpeg("fnRsQuality16S2023_test.jpg",width = 20, height =17, units = "in", res = 300)
plotQualityProfile(fnRs_test[1:45])
#truncate at 225

```

```{r}
#if needed, move samples with less than 1000 reads to cutadapt_removed folder and start again with only the samples with reads >1000

# CHANGE to the directory containing the trimmed unzipped fastq files.
path <- "/../../projects/gmgi/ecosystem-diversity/ossininglakes/Re-Run_2026/2023/16S/cutadapt"

#check all files are in the path
list.files(path)

```
```{r}
## Figuring out why out was getting rid of so manu reads originally
#plotQualityProfile(fnFs[1:2])
#plotQualityProfile(fnRs[1:2])
#fq <- ShortRead::readFastq(fnFs[1])
#summary(width(ShortRead::sread(fq)))
#fqR <- ShortRead::readFastq(fnRs[1])
#summary(width(ShortRead::sread(fqR)))

```

dada2
```{r}
# Forward and reverse fastq file names have format: SAMPLENAME_R1_trimmed.fastq and SAMPLENAME_R2_trimmed.fastq
fnFs <- sort(list.files(path, pattern="_R1_trimmed.fastq.gz", full.names = TRUE))
fnRs <- sort(list.files(path, pattern="_R2_trimmed.fastq.gz", full.names = TRUE))

# Extract sample names, assuming file names have format: SAMPLENAME_XXX.fastq
sample.names <- sapply(strsplit(basename(fnFs), "_"), `[`, 1)

# visualize the quality profiles of the forward reads (note where to truncate)
jpeg("fnFsQuality16S2023_trim.jpg",width = 20, height =17, units = "in", res = 300)
plotQualityProfile(fnFs[1:45])
#truncate at 230

# visualize the quality profiles of the reverse reads (note where to truncate)
jpeg("fnRsQuality16S2023_trim.jpg",width = 20, height =17, units = "in", res = 300)
plotQualityProfile(fnRs[1:45])
#truncate at 225

# Filter and trim
# Place filtered files in filtered subdirectory
filtFs <- file.path(path, "filtered", paste0(sample.names, "_F_filt.fastq.gz"))
filtRs <- file.path(path, "filtered", paste0(sample.names, "_R_filt.fastq.gz"))
names(filtFs) <- sample.names
names(filtRs) <- sample.names

# trim using standard filtering parameters: maxN=0 (DADA2 requires no Ns), truncQ=2, rm.phix=TRUE and maxEE=2. The maxEE parameter sets the maximum number of “expected errors” allowed in a read --> truncLen dependent on quality # On Windows set multithread=FALSE
out <- filterAndTrim(fnFs, filtFs, fnRs, filtRs, truncLen=c(230,225), maxN=0, maxEE=c(2,2), truncQ=2, rm.phix=TRUE, compress=TRUE, multithread=FALSE) 
head(out)

# learn error rates
errF <- learnErrors(filtFs, multithread=5)
errR <- learnErrors(filtRs, multithread=5)

# plot error rates (should be similar to theoretical)
jpeg("errF16S2023_trim.jpg",width = 10, height =8, units = "in", res = 300)
plotErrors(errF, nominalQ=TRUE)
jpeg("errR16S2023_trim.jpg",width = 10, height =8, units = "in", res = 300)
plotErrors(errR, nominalQ=TRUE)

# apply the core sample inference algorithm to forward and reverse reads
dadaFs <- dada(filtFs, err=errF, multithread=5)
dadaRs <- dada(filtRs, err=errR, multithread=5)

#Inspecting the returned dada-class object
dadaFs[[1]]

# merge paired reads
mergers <- mergePairs(dadaFs, filtFs, dadaRs, filtRs, verbose=TRUE)

# Inspect the merger data.frame from the first sample
head(mergers[[1]])

# build sequence table
seqtab <- makeSequenceTable(mergers)
dim(seqtab)

# Inspect distribution of sequence lengths
table(nchar(getSequences(seqtab)))

# remove chimeras
seqtab.nochim <- removeBimeraDenovo(seqtab, method="consensus", multithread=5, verbose=TRUE)
dim(seqtab.nochim)
sum(seqtab.nochim)/sum(seqtab)

# check number of reads that made it through the pipeline #many samples
getN <- function(x) sum(getUniques(x))
# for multiple samples:
track <- cbind(out, sapply(dadaFs, getN), sapply(dadaRs, getN), sapply(mergers, getN), rowSums(seqtab.nochim)) 
# single sample: If processing a single sample, remove the sapply calls: e.g. replace sapply(dadaFs, getN) with getN(dadaFs)
##track <- cbind(out, getN(dadaFs), getN(dadaRs), getN(mergers), rowSums(seqtab.nochim)) 

# check
colnames(track) <- c("input", "filtered", "denoisedF", "denoisedR", "merged", "nonchim")
rownames(track) <- sample.names
head(track)

# assign taxonomy
taxa <- assignTaxonomy(seqtab.nochim, "/../../projects/gmgi/databases/16S/CyanoSeq_1.2_SILVA138.1_dada2.fastq.gz", multithread=5)
taxa <- addSpecies(taxa, "/../../projects/gmgi/databases/16S/silva-138.1/silva_species_assignment_v138.1.fa.gz")

# inspect taxonomic assignments
taxa.print <- taxa #Removing sequence rownames for display only
rownames(taxa.print) <- NULL
head(taxa.print)

# merge taxonomy and counts
asv_seqs <- colnames(seqtab.nochim)
asv_headers <- vector(dim(seqtab.nochim)[2], mode="character")

# fasta header as ASV_
for (i in 1:dim(seqtab.nochim)[2]) {
  asv_headers[i] <- paste(">ASV", i, sep="_")
}

# ASV sequences
asv_fasta <- c(rbind(asv_headers, asv_seqs))

# ASV abundance
asv_otu <- t(seqtab.nochim)
row.names(asv_otu) <- sub(">", "", asv_headers)

# ASV taxonomy
asv_tax <- taxa
row.names(asv_tax) <- sub(">", "", asv_headers)

# merging abundance and tax table
OTU_TAX_table <- merge(asv_otu, asv_tax, by=0)

# writing output files and saving to server
write(asv_fasta, "/../../projects/gmgi/ecosystem-diversity/ossininglakes/Re-Run_2026/2023/16S/dada2/asv_fasta_rl.fa")
write.table(asv_otu, "/../../projects/gmgi/ecosystem-diversity/ossininglakes/Re-Run_2026/2023/16S/dada2/asv_otu_rl.csv", sep=",", quote=F, col.names=NA)
write.table(asv_tax, "/../../projects/gmgi/ecosystem-diversity/ossininglakes/Re-Run_2026/2023/16S/dada2/asv_tax_rl.csv", sep=",", quote=F, col.names=NA)
write.table(OTU_TAX_table, "/../../projects/gmgi/ecosystem-diversity/ossininglakes/Re-Run_2026/2023/16S/dada2/OTU_TAX_table_rl.csv", sep=",", quote=F, col.names=NA)
```

phyloseq (post dada2)
```{r}
#phyloseq object creation
ps <- phyloseq(otu_table(seqtab.nochim, taxa_are_rows=FALSE), 
               tax_table(taxa))
ps

#richness (alpha diversity)
jpeg("richness16S2023_trim.jpg",width = 10, height =8, units = "in", res = 300)
plot_richness(ps, measures=c("Observed"))

#ordinate
ps.prop <- transform_sample_counts(ps, function(otu) otu/sum(otu))
ord.nmds.bray <- ordinate(ps.prop, method="NMDS", distance="bray")
#jpeg("ordination16S2023_trim.jpg",width = 10, height =8, units = "in", res = 300)
#plot_ordination(ps.prop, ord.nmds.bray, color= , title="Bray NMDS")

# NMDS
os.ord <- ordinate(ps, "NMDS", "bray")


#Bar plot
top20 <- names(sort(taxa_sums(ps), decreasing=TRUE))[1:30]
ps.top20 <- transform_sample_counts(ps, function(OTU) OTU/sum(OTU))
ps.top20 <- prune_taxa(top20, ps.top20)
jpeg("barplot16S2023_trim.jpg",width = 14, height =8, units = "in", res = 300)
plot_bar(ps, fill="Phylum")
```