---
title: "Ossining Lakes Microbial Community Analysis (2023)"
author: "Taylor F. Gibson, Kayley Pugh, Matthew J. Harke"
date: "`r format(Sys.Date(), '%m/%d/%Y')`"
knit: (function(input_file, encoding) {
  out_dir <- 'docs';
  rmarkdown::render(input_file,
    encoding = encoding,
    output_file = file.path(dirname(input_file), out_dir, 'index.html'))})
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    number_sections: true
    theme: cosmo
    highlight: tango
    code_folding: show
    fig_width: 8
    fig_height: 6
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = TRUE,
  message = TRUE,
  fig.align = "center",
  fig.path = "figs/"
)
```

# Overview

**[Manuscript title here]**

Gibson, T.F., Pugh, K., & Harke, M.J. (*Submitted*) [Manuscript title]. *Harmful Algae*.

## Summary

Analysis of 16S (prokaryote) and 18S (eukaryote) amplicon sequence variant (ASV)
data from three Ossining-area lakes — Lake Mohegan (LM), Lake Rippowam (LR), and
Teatown Lake (TTL) — sampled from June to September 2023. Environmental metadata
(nutrients, temperature, pH, dissolved oxygen, microcystin) were measured at each
sampling event.

This document contains code for:

- Diversity analysis of 16S and 18S amplicon data (alpha diversity, community
  composition, PCoA ordination, PERMANOVA)
- Weighted Gene Co-expression Network Analysis (WGCNA) to identify modules of
  co-occurring ASVs
- Module–trait (environmental parameter) correlation analysis
- Intramodular connectivity and hub taxa identification
- Inter-domain (16S vs 18S) correlation analysis within modules
- Network visualization of inter-domain associations

## Contents

All input files are available in the `data/` directory of the
[GitHub repository](https://github.com/mharke/OssiningLakes).
Raw sequences are publicly available under
[BioProject PRJNA_XXXXXXX](https://www.ncbi.nlm.nih.gov/bioproject/PRJNA_XXXXXXX).

# Set up working R environment

The following analysis was performed in **R version `r getRversion()`**.

```{r load-libraries, message=FALSE, warning=FALSE}
# --- Core analysis packages ---
library(WGCNA)
library(vegan)
library(RColorBrewer)

# --- Phyloseq and microbiome ---
library(phyloseq)
library(microbiome)

# --- Tidyverse for data wrangling and plotting ---
library(tidyverse)    # includes ggplot2, dplyr, readr, tidyr, stringr, purrr
library(lubridate)
library(readxl)

# --- Network visualization ---
library(igraph)
library(ggraph)

# --- Plot layout ---
library(patchwork)

# --- Pairwise PERMANOVA ---
library(metagMisc)

# Enable multithreading for WGCNA
enableWGCNAThreads()
```

```{r utility-functions}
# Null-coalescing operator (used by explore_module_trait)
`%||%` <- function(a, b) if (is.null(a) || length(a) == 0) b else a

# Create output directory for WGCNA results
dir.create("results/wgcna", recursive = TRUE, showWarnings = FALSE)
```

# 16S and 18S Diversity Analysis (2023)

## 16S Prokaryote Community (2023) {.tabset}

> **NOTE:** This section requires the DADA2-derived ASV tables (`asv_otu_rl.csv`,
> `asv_tax_rl.csv`) for 16S 2023 data and the sample metadata with `Name`,
> `Date_Sam`, `Lake`, and `Month` columns. These files are generated upstream by
> the DADA2 pipeline and are not included in this repository. If you do not have
> these files, set `eval = FALSE` for these chunks or skip to [WGCNA Analysis].
> Raw sequences are available under BioProject PRJNA_XXXXXXX.

### Import data and create phyloseq object

```{r import-16S-2023, eval=FALSE}
# --- Import 16S 2023 data ---
# Sequence count table (DADA2 output)
otutab_16S23 <- read.csv("data/16S_2023_asv_otu_rl.csv", check.names = FALSE)
rownames(otutab_16S23) <- otutab_16S23[, 1]
otutab_16S23 <- otutab_16S23[, 2:46]
otutab_16S23 <- otu_table(otutab_16S23, taxa_are_rows = TRUE)

# Sequence taxonomy classifications (DADA2 output)
tax_16S23 <- read.csv("data/16S_2023_asv_tax_rl.csv", header = TRUE)
rownames(tax_16S23) <- taxa_names(otutab_16S23)
tax_16S23 <- tax_16S23[, 2:9]
tax_16S23 <- as.matrix(tax_16S23)
tax_16S23 <- tax_table(tax_16S23)
taxa_names(tax_16S23) <- taxa_names(otutab_16S23)

# Sample metadata
meta_16S23 <- read_xlsx("data/Ossining_Lake_Metadata_2023.xlsx")
meta_16S23 <- sample_data(meta_16S23)
rownames(meta_16S23) <- meta_16S23$Name

# Build phyloseq object
physeq_16S23 <- merge_phyloseq(otutab_16S23, tax_16S23, meta_16S23)

# Remove samples with zero counts
physeq_16S23 <- prune_samples(sample_sums(physeq_16S23) >= 1, physeq_16S23)

# Remove chloroplast sequences
physeq_16S23 <- subset_taxa(physeq_16S23, Order != "Chloroplast")
```

### Alpha diversity

```{r alpha-16S-2023, eval=FALSE}
# Calculate alpha diversity metrics
alpha_16S23 <- microbiome::alpha(physeq_16S23)
write.csv(alpha_16S23, "results/16S_2023_alpha_diversity.csv")

# Plot observed richness by sampling date, faceted by lake
plot_richness(physeq_16S23, x = "Date_Sam", measures = "Observed",
              title = "16S 2023 Richness") +
  geom_boxplot() +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 60, hjust = 1),
        axis.text = element_text(size = 14),
        axis.title.x = element_blank()) +
  facet_grid(. ~ Lake, scales = "free")
```

### Phylum-level community composition

```{r phylum-barplot-16S-2023, eval=FALSE}
# Agglomerate to phylum level, convert to relative abundance
p_16S23 <- subset_taxa(physeq_16S23) %>%
  tax_glom(taxrank = "Phylum") %>%
  transform_sample_counts(function(x) x / sum(x)) %>%
  psmelt()

# Summarize relative abundance by lake, month, and phylum
relabund_16S23_phylum <- p_16S23 %>%
  group_by(Lake, Month, Date_Sam, Phylum) %>%
  summarize(RelAbund = sum(Abundance), .groups = "drop")
write.csv(relabund_16S23_phylum,
          "results/16S_2023_relative_abundance_phylum.csv",
          row.names = FALSE)

# Generate color palette
colourCount <- length(unique(p_16S23$Phylum))
getPalette <- colorRampPalette(brewer.pal(12, "Paired"))

# Plot phylum-level relative abundance by lake
ggplot(arrange(p_16S23, Phylum),
       aes(x = Date_Sam, y = Abundance, fill = Phylum)) +
  geom_bar(stat = "identity") +
  theme_bw() +
  theme(axis.title.x = element_blank(),
        plot.title = element_text(size = 10, hjust = 0.5),
        legend.text = element_text(size = 10),
        axis.text = element_text(size = 10),
        axis.text.x = element_text(angle = 45, hjust = 1),
        strip.text.x = element_text(size = 9)) +
  ylab("Relative Read Abundance") +
  scale_fill_manual(values = getPalette(colourCount)) +
  scale_y_continuous(limits = c(0, 1), expand = c(0, 0)) +
  guides(fill = guide_legend(ncol = 2)) +
  facet_grid(. ~ Lake, scales = "free") +
  ggtitle("16S 2023 Relative Read Abundance")
```

### Cyanobacteria community composition

```{r cyano-barplot-16S-2023, eval=FALSE}
# Subset Cyanobacteriota, agglomerate to edited genus level (Taxa column)
cyano_16S23 <- subset_taxa(physeq_16S23, Phylum == "Cyanobacteriota") %>%
  tax_glom(taxrank = "Taxa") %>%
  transform_sample_counts(function(x) x / sum(x)) %>%
  psmelt() %>%
  filter(Abundance > 0.01)

# Summarize relative abundance
relabund_16S23_cyano <- cyano_16S23 %>%
  group_by(Lake, Month, Date_Sam, Taxa) %>%
  summarize(RelAbund = sum(Abundance), .groups = "drop")
write.csv(relabund_16S23_cyano,
          "results/16S_2023_relative_abundance_cyano.csv",
          row.names = FALSE)

# Generate color palette
colourCount <- length(unique(cyano_16S23$Taxa))
getPalette <- colorRampPalette(brewer.pal(8, "Set1"))

# Plot cyanobacteria relative abundance by lake
ggplot(arrange(cyano_16S23, Taxa),
       aes(x = Date_Sam, y = Abundance, fill = Taxa)) +
  geom_bar(stat = "identity") +
  theme_bw() +
  theme(plot.title = element_text(size = 10, hjust = 0.5),
        axis.title.x = element_blank(),
        legend.text = element_text(size = 10),
        legend.position = "right",
        axis.text = element_text(size = 10),
        axis.text.x = element_text(angle = 45, hjust = 1),
        strip.text.x = element_text(size = 9)) +
  ylab("Relative Read Abundance") +
  scale_fill_manual(values = getPalette(colourCount)) +
  scale_y_continuous(limits = c(0, 1.001), expand = c(0, 0)) +
  guides(fill = guide_legend(ncol = 1)) +
  facet_grid(. ~ Lake, scales = "free") +
  ggtitle("16S 2023 Cyanobacteria Relative Read Abundance")
```

### PCoA ordination

```{r pcoa-16S-2023, eval=FALSE}
# --- PCoA for each lake individually ---

# Lake Mohegan
LM_16S23 <- subset_samples(physeq_16S23, Lake == "Lake Mohegan")
LM_16S23rare <- rarefy_even_depth(LM_16S23)
pcoa_LM_16S23 <- ordinate(physeq = LM_16S23rare, method = "PCoA", distance = "bray")
ord_LM_16S23 <- plot_ordination(physeq = LM_16S23rare,
                                ordination = pcoa_LM_16S23,
                                color = "Month",
                                title = "Lake Mohegan") +
  geom_point(aes(color = Month), alpha = 0.75, size = 4) +
  theme_bw() +
  theme(text = element_text(size = 14)) +
  guides(color = guide_legend(title = "Month", ncol = 1))

# Lake Rippowam
LR_16S23 <- subset_samples(physeq_16S23, Lake == "Lake Rippowam")
LR_16S23rare <- rarefy_even_depth(LR_16S23)
pcoa_LR_16S23 <- ordinate(physeq = LR_16S23rare, method = "PCoA", distance = "bray")
ord_LR_16S23 <- plot_ordination(physeq = LR_16S23rare,
                                ordination = pcoa_LR_16S23,
                                color = "Month",
                                title = "Lake Rippowam") +
  geom_point(aes(color = Month), alpha = 0.75, size = 4) +
  theme_bw() +
  theme(text = element_text(size = 14)) +
  guides(color = guide_legend(title = "Month", ncol = 1))

# Teatown Lake
TTL_16S23 <- subset_samples(physeq_16S23, Lake == "Teatown Lake")
TTL_16S23rare <- rarefy_even_depth(TTL_16S23)
pcoa_TTL_16S23 <- ordinate(physeq = TTL_16S23rare, method = "PCoA", distance = "bray")
ord_TTL_16S23 <- plot_ordination(physeq = TTL_16S23rare,
                                 ordination = pcoa_TTL_16S23,
                                 color = "Month",
                                 title = "Teatown Lake") +
  geom_point(aes(color = Month), alpha = 0.75, size = 4) +
  theme_bw() +
  theme(text = element_text(size = 14)) +
  guides(color = guide_legend(title = "Month", ncol = 1))

# Combined PCoA plot (patchwork)
ord_LM_16S23 / ord_LR_16S23 / ord_TTL_16S23 +
  plot_annotation("PCoA of 16S 2023 Data")
```

### Cyanobacteria PCoA

```{r pcoa-cyano-16S-2023, eval=FALSE}
# Subset cyanobacteria for PCoA
cb_16S23 <- subset_taxa(physeq_16S23, Phylum == "Cyanobacteriota")

# Lake Mohegan cyanobacteria
LMcb_16S23 <- subset_samples(cb_16S23, Lake == "Lake Mohegan")
LMcbrare_16S23 <- rarefy_even_depth(LMcb_16S23)
LMcbpca_16S23 <- ordinate(LMcbrare_16S23, "PCoA")
cb_ord_LM_16S23 <- plot_ordination(LMcbrare_16S23, LMcbpca_16S23,
                                    color = "Month",
                                    title = "Lake Mohegan") +
  geom_point(aes(color = Month), alpha = 0.75, size = 4) +
  theme_bw() +
  theme(text = element_text(size = 14)) +
  guides(color = guide_legend(title = "Month", ncol = 1))

# Lake Rippowam cyanobacteria
LRcb_16S23 <- subset_samples(cb_16S23, Lake == "Lake Rippowam")
LRcbrare_16S23 <- rarefy_even_depth(LRcb_16S23)
LRcbpca_16S23 <- ordinate(LRcbrare_16S23, "PCoA")
cb_ord_LR_16S23 <- plot_ordination(LRcbrare_16S23, LRcbpca_16S23,
                                    color = "Month",
                                    title = "Lake Rippowam") +
  geom_point(aes(color = Month), alpha = 0.75, size = 4) +
  theme_bw() +
  theme(text = element_text(size = 14)) +
  guides(color = guide_legend(title = "Month", ncol = 1))

# Teatown Lake cyanobacteria
TTLcb_16S23 <- subset_samples(cb_16S23, Lake == "Teatown Lake")
TTLcbrare_16S23 <- rarefy_even_depth(TTLcb_16S23)
TTLcbpca_16S23 <- ordinate(TTLcbrare_16S23, "PCoA")
cb_ord_TTL_16S23 <- plot_ordination(TTLcbrare_16S23, TTLcbpca_16S23,
                                     color = "Month",
                                     title = "Teatown Lake") +
  geom_point(aes(color = Month), alpha = 0.75, size = 4) +
  theme_bw() +
  theme(text = element_text(size = 14)) +
  guides(color = guide_legend(title = "Month", ncol = 1))

# Combined cyanobacteria PCoA
cb_ord_LM_16S23 / cb_ord_LR_16S23 / cb_ord_TTL_16S23 +
  plot_annotation("Cyanobacteria PCoA of 16S 2023 Data")
```

### PERMANOVA

```{r permanova-16S-2023, eval=FALSE}
# --- Whole community PERMANOVA by lake ---

# Lake Mohegan
LMbray16S23 <- phyloseq::distance(LM_16S23rare, method = "bray")
LMdf16S23 <- data.frame(sample_data(LM_16S23rare))
LMperm16S23 <- adonis2(LMbray16S23 ~ Month, data = LMdf16S23)
LMperm16S23  # significant

# Lake Rippowam
LRbray16S23 <- phyloseq::distance(LR_16S23rare, method = "bray")
LRdf16S23 <- data.frame(sample_data(LR_16S23rare))
LRperm16S23 <- adonis2(LRbray16S23 ~ Month, data = LRdf16S23)
LRperm16S23  # significant

# Teatown Lake
TTLbray16S23 <- phyloseq::distance(TTL_16S23rare, method = "bray")
TTLdf16S23 <- data.frame(sample_data(TTL_16S23rare))
TTLperm16S23 <- adonis2(TTLbray16S23 ~ Month, data = TTLdf16S23)
TTLperm16S23  # significant

# --- Cyanobacteria PERMANOVA by lake ---

cLMbray16S23 <- phyloseq::distance(LMcbrare_16S23, method = "bray")
cLMdf16S23 <- data.frame(sample_data(LMcbrare_16S23))
cLMperm16S23 <- adonis2(cLMbray16S23 ~ Month, data = cLMdf16S23)
cLMperm16S23  # significant

cLRbray16S23 <- phyloseq::distance(LRcbrare_16S23, method = "bray")
cLRdf16S23 <- data.frame(sample_data(LRcbrare_16S23))
cLRperm16S23 <- adonis2(cLRbray16S23 ~ Month, data = cLRdf16S23)
cLRperm16S23  # significant

cTTLbray16S23 <- phyloseq::distance(TTLcbrare_16S23, method = "bray")
cTTLdf16S23 <- data.frame(sample_data(TTLcbrare_16S23))
cTTLperm16S23 <- adonis2(cTTLbray16S23 ~ Month, data = cTTLdf16S23)
cTTLperm16S23  # significant

# --- All lakes combined PERMANOVA (Lake * Month interaction) ---

# Whole community
Lake23rare16 <- rarefy_even_depth(physeq_16S23)
Lake23bray16 <- phyloseq::distance(Lake23rare16, method = "bray")
Lake23df16 <- data.frame(sample_data(Lake23rare16))
Lake23perm16 <- adonis2(Lake23bray16 ~ Lake * Month, data = Lake23df16)
Lake23perm16  # All combinations significant

# Cyanobacteria only
cb_16S23rare <- rarefy_even_depth(cb_16S23)
cLake23bray <- phyloseq::distance(cb_16S23rare, method = "bray")
cLake23df <- data.frame(sample_data(cb_16S23rare))
cLake23perm <- adonis2(cLake23bray ~ Lake * Month, data = cLake23df)
cLake23perm  # All combinations significant
```

### Pairwise PERMANOVA

```{r pairwise-permanova-16S-2023, eval=FALSE}
# Pairwise comparisons following significant PERMANOVA results

# Whole community: per-lake pairwise by Month
LMdf16S23$Month <- as.factor(LMdf16S23$Month)
LMpairwise16S23 <- adonis_pairwise(x = LMdf16S23, dd = LMbray16S23,
                                    group.var = "Month")
LMpairwise16S23  # all month combos significant

LRdf16S23$Month <- as.factor(LRdf16S23$Month)
LRpairwise16S23 <- adonis_pairwise(x = LRdf16S23, dd = LRbray16S23,
                                    group.var = "Month")
LRpairwise16S23  # all significant except Aug/Sept

TTLdf16S23$Month <- as.factor(TTLdf16S23$Month)
TTLpairwise16S23 <- adonis_pairwise(x = TTLdf16S23, dd = TTLbray16S23,
                                     group.var = "Month")
TTLpairwise16S23  # all significant except Aug/Sept, July/June

# Cyanobacteria: per-lake pairwise by Month
cLMdf16S23$Month <- as.factor(cLMdf16S23$Month)
cLMpairwise16S23 <- adonis_pairwise(x = cLMdf16S23, dd = cLMbray16S23,
                                     group.var = "Month")
cLMpairwise16S23  # all significant except July/June

cLRdf16S23$Month <- as.factor(cLRdf16S23$Month)
cLRpairwise16S23 <- adonis_pairwise(x = cLRdf16S23, dd = cLRbray16S23,
                                     group.var = "Month")
cLRpairwise16S23  # Aug/July, Aug/June, July/June, June/Sept significant

cTTLdf16S23$Month <- as.factor(cTTLdf16S23$Month)
cTTLpairwise16S23 <- adonis_pairwise(x = cTTLdf16S23, dd = cTTLbray16S23,
                                      group.var = "Month")
cTTLpairwise16S23  # Aug/July, Aug/June, July/Sept, June/Sept significant

# All lakes combined: pairwise by Lake and by Month
Lake23df16$Month <- as.factor(Lake23df16$Month)
Lake23df16$Lake <- as.factor(Lake23df16$Lake)

Lake23pairwise16 <- adonis_pairwise(x = Lake23df16, dd = Lake23bray16,
                                     group.var = "Lake")
Lake23pairwise16  # all 3 lake combos significant

Month23pairwise16 <- adonis_pairwise(x = Lake23df16, dd = Lake23bray16,
                                      group.var = "Month")
Month23pairwise16  # all significant except Aug/Sept

# All lakes cyanobacteria combined
cLake23df$Month <- as.factor(cLake23df$Month)
cLake23df$Lake <- as.factor(cLake23df$Lake)

cLake23pairwise <- adonis_pairwise(x = cLake23df, dd = cLake23bray,
                                    group.var = "Lake")
cLake23pairwise  # all 3 lake combos significant

cMonth23pairwise <- adonis_pairwise(x = cLake23df, dd = cLake23bray,
                                     group.var = "Month")
cMonth23pairwise  # all significant except July/June
```

## 18S Eukaryote Community (2023) {.tabset}

> **NOTE:** This section requires the DADA2-derived ASV tables for 18S 2023 data
> (PR2 reference database). These are generated upstream and are not included in
> this repository. Set `eval = FALSE` if input files are unavailable.

### Import data and create phyloseq object

```{r import-18S-2023, eval=FALSE}
# --- Import 18S 2023 data ---
# Sequence count table (DADA2 output, PR2 database)
otutab_18S23 <- read.csv("data/18S_2023_asv_otu_rl.csv", check.names = FALSE)
rownames(otutab_18S23) <- otutab_18S23[, 1]
otutab_18S23 <- otutab_18S23[, 2:46]
otutab_18S23 <- otu_table(otutab_18S23, taxa_are_rows = TRUE)

# Sequence taxonomy classifications (PR2 database)
tax_18S23 <- read.csv("data/18S_2023_asv_tax_rl.csv", header = TRUE)
rownames(tax_18S23) <- taxa_names(otutab_18S23)
tax_18S23 <- tax_18S23[, 2:11]
tax_18S23 <- as.matrix(tax_18S23)
tax_18S23 <- tax_table(tax_18S23)
taxa_names(tax_18S23) <- taxa_names(otutab_18S23)

# Sample metadata
meta_18S23 <- read_xlsx("data/Ossining_Lake_Metadata_2023.xlsx")
meta_18S23 <- sample_data(meta_18S23)
#rownames(meta_18S23) <- meta_18S23$Name2

# Build phyloseq object
physeq_18S23 <- merge_phyloseq(otutab_18S23, tax_18S23, meta_18S23)

# Remove samples with zero counts
physeq_18S23 <- prune_samples(sample_sums(physeq_18S23) >= 1, physeq_18S23)
```

### Alpha diversity

```{r alpha-18S-2023, eval=FALSE}
# Calculate alpha diversity metrics
alpha_18S23 <- microbiome::alpha(physeq_18S23)
write.csv(alpha_18S23, "results/18S_2023_alpha_diversity.csv")

# Plot observed richness by sampling date, faceted by lake
plot_richness(physeq_18S23, x = "Date_Sam", measures = "Observed",
              title = "18S 2023 Richness") +
  geom_boxplot() +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 60, hjust = 1),
        axis.text = element_text(size = 14),
        axis.title.x = element_blank()) +
  facet_grid(. ~ Lake, scales = "free")
```

### Division-level community composition

```{r division-barplot-18S-2023, eval=FALSE}
# Agglomerate to edited division level (Taxa column), convert to relative abundance
p_18S23 <- subset_taxa(physeq_18S23) %>%
  tax_glom(taxrank = "Taxa") %>%
  transform_sample_counts(function(x) x / sum(x)) %>%
  psmelt()

# Summarize relative abundance
relabund_18S23_div <- p_18S23 %>%
  group_by(Lake, Month, Date_Sam, Taxa) %>%
  summarize(RelAbund = sum(Abundance), .groups = "drop")
write.csv(relabund_18S23_div,
          "results/18S_2023_relative_abundance_division.csv",
          row.names = FALSE)

# Generate color palette
colourCount <- length(unique(p_18S23$Taxa))
getPalette <- colorRampPalette(brewer.pal(12, "Paired"))

# Plot division-level relative abundance by lake
ggplot(arrange(p_18S23, Taxa),
       aes(x = Date_Sam, y = Abundance, fill = Taxa)) +
  geom_bar(stat = "identity") +
  theme_bw() +
  theme(axis.title.x = element_blank(),
        plot.title = element_text(size = 10, hjust = 0.5),
        legend.text = element_text(size = 10),
        axis.text = element_text(size = 10),
        axis.text.x = element_text(angle = 45, hjust = 1),
        strip.text.x = element_text(size = 9)) +
  ylab("Relative Read Abundance") +
  scale_fill_manual(values = getPalette(colourCount)) +
  scale_y_continuous(limits = c(0, 1), expand = c(0, 0)) +
  guides(fill = guide_legend(ncol = 2)) +
  facet_grid(. ~ Lake, scales = "free") +
  ggtitle("18S 2023 Relative Read Abundance")
```

### Chytrid community composition

```{r chytrid-barplot-18S-2023, eval=FALSE}
# Subset Chytridiomycota, agglomerate to order level
chytrid_18S23 <- subset_taxa(physeq_18S23, Class == "Chytridiomycota") %>%
  tax_glom(taxrank = "Order") %>%
  transform_sample_counts(function(x) x / sum(x)) %>%
  psmelt() %>%
  filter(Abundance > 0.01)

# Summarize relative abundance
relabund_18S23_chytrid <- chytrid_18S23 %>%
  group_by(Lake, Month, Date_Sam, Order) %>%
  summarize(RelAbund = sum(Abundance), .groups = "drop")
write.csv(relabund_18S23_chytrid,
          "results/18S_2023_relative_abundance_chytrid.csv",
          row.names = FALSE)

# Generate color palette
colourCount <- length(unique(chytrid_18S23$Order))
getPalette <- colorRampPalette(brewer.pal(8, "Set1"))

# Plot chytrid relative abundance by lake
ggplot(arrange(chytrid_18S23, Order),
       aes(x = Date_Sam, y = Abundance, fill = Order)) +
  geom_bar(stat = "identity") +
  theme_bw() +
  theme(plot.title = element_text(size = 10, hjust = 0.5),
        axis.title.x = element_blank(),
        legend.text = element_text(size = 10),
        legend.position = "right",
        axis.text = element_text(size = 10),
        axis.text.x = element_text(angle = 45, hjust = 1),
        strip.text.x = element_text(size = 9)) +
  ylab("Relative Read Abundance") +
  scale_fill_manual(values = getPalette(colourCount)) +
  scale_y_continuous(limits = c(0, 1.001), expand = c(0, 0)) +
  guides(fill = guide_legend(ncol = 1)) +
  facet_grid(. ~ Lake, scales = "free") +
  ggtitle("18S 2023 Chytrid Relative Read Abundance")
```

### Metazoa community composition

```{r metazoa-barplot-18S-2023, eval=FALSE}
# Subset Metazoa, agglomerate to genus level
metazoa_18S23 <- subset_taxa(physeq_18S23, Subdivision == "Metazoa") %>%
  tax_glom(taxrank = "Genus") %>%
  transform_sample_counts(function(x) x / sum(x)) %>%
  psmelt()

# Summarize relative abundance
relabund_18S23_metazoa <- metazoa_18S23 %>%
  group_by(Lake, Month, Date_Sam, Genus) %>%
  summarize(RelAbund = sum(Abundance), .groups = "drop")
write.csv(relabund_18S23_metazoa,
          "results/18S_2023_relative_abundance_metazoa.csv",
          row.names = FALSE)

# Generate color palette
colourCount <- length(unique(metazoa_18S23$Genus))
getPalette <- colorRampPalette(brewer.pal(8, "Set1"))

# Plot metazoa relative abundance by lake
ggplot(arrange(metazoa_18S23, Genus),
       aes(x = Date_Sam, y = Abundance, fill = Genus)) +
  geom_bar(stat = "identity") +
  theme_bw() +
  theme(plot.title = element_text(size = 10, hjust = 0.5),
        axis.title.x = element_blank(),
        legend.text = element_text(size = 10),
        legend.position = "right",
        axis.text = element_text(size = 10),
        axis.text.x = element_text(angle = 45, hjust = 1),
        strip.text.x = element_text(size = 9)) +
  ylab("Relative Read Abundance") +
  scale_fill_manual(values = getPalette(colourCount)) +
  scale_y_continuous(limits = c(0, 1.001), expand = c(0, 0)) +
  guides(fill = guide_legend(ncol = 2)) +
  facet_grid(. ~ Lake, scales = "free") +
  ggtitle("18S 2023 Metazoa Relative Read Abundance")
```

### PCoA ordination

```{r pcoa-18S-2023, eval=FALSE}
# --- PCoA for each lake individually ---

# Lake Mohegan
LM_18S23 <- subset_samples(physeq_18S23, Lake == "Lake Mohegan")
LM_18S23rare <- rarefy_even_depth(LM_18S23)
pcoa_LM_18S23 <- ordinate(physeq = LM_18S23rare, method = "PCoA", distance = "bray")
ord_LM_18S23 <- plot_ordination(physeq = LM_18S23rare,
                                ordination = pcoa_LM_18S23,
                                color = "Month",
                                title = "Lake Mohegan") +
  geom_point(aes(color = Month), alpha = 0.75, size = 4) +
  theme_bw() +
  theme(text = element_text(size = 14)) +
  guides(color = guide_legend(title = "Month", ncol = 1))

# Lake Rippowam
LR_18S23 <- subset_samples(physeq_18S23, Lake == "Lake Rippowam")
LR_18S23rare <- rarefy_even_depth(LR_18S23)
pcoa_LR_18S23 <- ordinate(physeq = LR_18S23rare, method = "PCoA", distance = "bray")
ord_LR_18S23 <- plot_ordination(physeq = LR_18S23rare,
                                ordination = pcoa_LR_18S23,
                                color = "Month",
                                title = "Lake Rippowam") +
  geom_point(aes(color = Month), alpha = 0.75, size = 4) +
  theme_bw() +
  theme(text = element_text(size = 14)) +
  guides(color = guide_legend(title = "Month", ncol = 1))

# Teatown Lake
TTL_18S23 <- subset_samples(physeq_18S23, Lake == "Teatown Lake")
TTL_18S23rare <- rarefy_even_depth(TTL_18S23)
pcoa_TTL_18S23 <- ordinate(physeq = TTL_18S23rare, method = "PCoA", distance = "bray")
ord_TTL_18S23 <- plot_ordination(physeq = TTL_18S23rare,
                                 ordination = pcoa_TTL_18S23,
                                 color = "Month",
                                 title = "Teatown Lake") +
  geom_point(aes(color = Month), alpha = 0.75, size = 4) +
  theme_bw() +
  theme(text = element_text(size = 14)) +
  guides(color = guide_legend(title = "Month", ncol = 1))

# Combined PCoA plot
ord_LM_18S23 / ord_LR_18S23 / ord_TTL_18S23 +
  plot_annotation("PCoA of 18S 2023 Data")
```

### PERMANOVA

```{r permanova-18S-2023, eval=FALSE}
# --- 18S PERMANOVA ---

# Lake Mohegan
LMbray18S23 <- phyloseq::distance(LM_18S23rare, method = "bray")
LMdf18S23 <- data.frame(sample_data(LM_18S23rare))
LMperm18S23 <- adonis2(LMbray18S23 ~ Month, data = LMdf18S23)
LMperm18S23  # NOT significant

# Lake Rippowam
LRbray18S23 <- phyloseq::distance(LR_18S23rare, method = "bray")
LRdf18S23 <- data.frame(sample_data(LR_18S23rare))
LRperm18S23 <- adonis2(LRbray18S23 ~ Month, data = LRdf18S23)
LRperm18S23  # significant

# Teatown Lake
TTLbray18S23 <- phyloseq::distance(TTL_18S23rare, method = "bray")
TTLdf18S23 <- data.frame(sample_data(TTL_18S23rare))
TTLperm18S23 <- adonis2(TTLbray18S23 ~ Month, data = TTLdf18S23)
TTLperm18S23  # significant

# All lakes combined (Lake * Month)
Lake23rare18 <- rarefy_even_depth(physeq_18S23)
Lake23bray18 <- phyloseq::distance(Lake23rare18, method = "bray")
Lake23df18 <- data.frame(sample_data(Lake23rare18))
Lake23perm18 <- adonis2(Lake23bray18 ~ Lake * Month, data = Lake23df18)
Lake23perm18  # All combinations significant
```

### Pairwise PERMANOVA

```{r pairwise-permanova-18S-2023, eval=FALSE}
# Pairwise comparisons following significant 18S PERMANOVA results

# Lake Rippowam (significant)
LRdf18S23$Month <- as.factor(LRdf18S23$Month)
LRpairwise18S23 <- adonis_pairwise(x = LRdf18S23, dd = LRbray18S23,
                                    group.var = "Month")
LRpairwise18S23  # Aug/June, Aug/Sept, July/June, July/Sept, June/Sept significant

# Teatown Lake (significant)
TTLdf18S23$Month <- as.factor(TTLdf18S23$Month)
TTLpairwise18S23 <- adonis_pairwise(x = TTLdf18S23, dd = TTLbray18S23,
                                     group.var = "Month")
TTLpairwise18S23  # Aug/July, July/Sept, July/June significant

# All lakes combined
Lake23df18$Month <- as.factor(Lake23df18$Month)
Lake23df18$Lake <- as.factor(Lake23df18$Lake)

Lake23pairwise18 <- adonis_pairwise(x = Lake23df18, dd = Lake23bray18,
                                     group.var = "Lake")
Lake23pairwise18  # LM/LR, LR/TTL significant

Month23pairwise18 <- adonis_pairwise(x = Lake23df18, dd = Lake23bray18,
                                      group.var = "Month")
Month23pairwise18  # June/Sept, July/Sept, Aug/June, Aug/July significant
```


# WGCNA Analysis

## Data loading, transformation, and filtering

We read the combined ASV count + taxonomy Table SX files for 16S and 18S. The
`process_counts_from_table_sx()` function separates count columns from taxonomy
columns, prefixes ASV IDs with the domain tag to avoid collisions, applies a
Hellinger transformation, and filters low-abundance ASVs.

```{r define-processing-function}
# Function to read combined count+taxonomy Table SX files.
# Separates count columns from taxonomy columns, prefixes ASV IDs with the
# domain tag (e.g., "16S_") to avoid ID collisions when 16S and 18S are merged,
# and strips the domain prefix from sample column names so that samples align
# across domains and with the trait data.
#
# Parameters:
#   file_path   : path to a Table SX CSV (ASV IDs in col 1, counts, then taxonomy)
#   domain_tag  : character prefix for the domain, e.g. "16S" or "18S"
#   taxa_cols   : character vector of taxonomy column names to exclude from counts
#   filter_mean : Hellinger-transformed row-mean threshold for ASV retention
process_counts_from_table_sx <- function(file_path,
                                         domain_tag,
                                         taxa_cols,
                                         filter_mean = 0.05) {
  
  # Read file; first column (unnamed) contains ASV IDs -> row names
  raw <- read.csv(file = file_path, header = TRUE, row.names = 1,
                  check.names = FALSE)
  
  # Separate count columns from taxonomy columns
  count_cols <- setdiff(colnames(raw), taxa_cols)
  counts <- raw[, count_cols, drop = FALSE]
  
  # Strip domain prefix from sample names (e.g., "16S-LM-6-1" -> "LM-6-1")
  colnames(counts) <- sub(paste0("^", domain_tag, "-"), "", colnames(counts))
  
  # Prefix ASV IDs with domain tag to ensure uniqueness after rbind
  rownames(counts) <- paste0(domain_tag, "_", rownames(counts))
  
  # Hellinger transformation (vegan::decostand)
  counts_H <- decostand(counts, method = "hellinger")
  
  # Filter low-abundance ASVs based on row mean of transformed values
  rowmean <- apply(counts_H, 1, mean)
  to_include <- rowmean >= filter_mean
  
  message(sprintf("[%s] ASVs kept: %d / %d", domain_tag,
                  sum(to_include), nrow(counts_H)))
  
  return(counts_H[to_include, ])
}
```

```{r load-and-filter-counts}
# Define taxonomy column names for each domain
taxa_cols_16s <- c("Kingdom", "Phylum", "Class", "Order",
                   "Family", "Genus", "Species", "Taxa")
taxa_cols_18s <- c("Domain", "Supergroup", "Division", "Subdivision",
                   "Class", "Order", "Family", "Genus", "Species", "Taxa")

# Process counts from the Table SX files
counts_filt16s <- process_counts_from_table_sx(
  file_path  = "data/Table_SX_16S_2023_ASV_Count_and_Taxonomy.csv",
  domain_tag = "16S",
  taxa_cols  = taxa_cols_16s
)
counts_filt18s <- process_counts_from_table_sx(
  file_path  = "data/Table_SX_18S_2023_ASV_Count_and_Taxonomy.csv",
  domain_tag = "18S",
  taxa_cols  = taxa_cols_18s
)

# Merge 16S and 18S (ASVs as rows, samples as columns), then transpose
# for WGCNA (samples as rows, ASVs as columns)
counts <- rbind(counts_filt16s, counts_filt18s)
tcounts <- as.data.frame(t(counts))

# Check for excessive missing data or low-variance genes
gsg <- goodSamplesGenes(tcounts, verbose = 3)
if (!gsg$allOK) {
  warning("Some samples or genes flagged by goodSamplesGenes. Review gsg$goodSamples and gsg$goodGenes.")
  tcounts <- tcounts[gsg$goodSamples, gsg$goodGenes]
}
```

## Environmental trait data preparation

Environmental metadata (nutrients, temperature, pH, dissolved oxygen, microcystin)
are matched to the sample order in the transposed count matrix.

```{r prepare-trait-data}
# Read trait (environmental) data
traitData <- read.csv("data/Ossining_Lake_Metadata_2023.csv", header = TRUE)

# The Site column uses underscores (e.g., "LR_6_1") but the Table SX sample
# names use hyphens (e.g., "LR-6-1"). Convert to match.
traitData$Site <- gsub("_", "-", traitData$Site)

# Match trait rows to sample order in tcounts
OTUSamples <- rownames(tcounts)
traitRows <- match(OTUSamples, traitData$Site)

# Verify all samples matched
if (any(is.na(traitRows))) {
  warning(paste("Unmatched samples:",
                paste(OTUSamples[is.na(traitRows)], collapse = ", ")))
}

# Drop both Site (col 1) AND SampleID (col 2) — only keep numeric traits
datTraits <- traitData[traitRows, -c(1, 2)]
rownames(datTraits) <- traitData[traitRows, 1]

# Report any missing data in trait columns
na_counts <- colSums(is.na(datTraits))
if (any(na_counts > 0)) {
  message("WARNING: Missing data detected in trait columns:")
  print(na_counts[na_counts > 0])
  message("Pairwise-complete correlations will have reduced N for these traits.")
}

collectGarbage()
```

## Sample clustering and outlier detection

Hierarchical clustering of samples to check for outliers prior to network
construction.

```{r sample-clustering, fig.width=12, fig.height=9}
# Sample clustering dendrogram
sampleTree <- hclust(dist(tcounts), method = "average")
par(cex = 0.6, mar = c(0, 4, 2, 0))
plot(sampleTree, main = "Sample clustering to detect outliers",
     sub = "", xlab = "", cex.lab = 1.5, cex.axis = 1.5, cex.main = 2)
```

```{r dendro-trait-heatmap, fig.width=12, fig.height=9}
# Visualize traits alongside sample dendrogram
sampleTree2 <- hclust(dist(tcounts), method = "average")
traitColors <- numbers2colors(datTraits, signed = FALSE)
plotDendroAndColors(sampleTree2, traitColors,
                    groupLabels = names(datTraits),
                    main = "Sample dendrogram and trait heatmap")
```

## Soft threshold selection

Pick the soft-thresholding power for a signed network. We look for the lowest
power where scale-free topology fit (R²) exceeds 0.70.

```{r pick-soft-threshold, fig.width=12, fig.height=5}
## Figure S2
powers <- c(1:10, seq(from = 12, to = 20, by = 2))
sft <- pickSoftThreshold(tcounts, powerVector = powers, verbose = 5,
                         networkType = "signed")

# Plot scale-free topology fit and mean connectivity
par(mfrow = c(1, 2), mar = c(5, 5, 3, 1))
cex1 <- 0.9

# Scale independence plot
plot(sft$fitIndices[, 1],
     -sign(sft$fitIndices[, 3]) * sft$fitIndices[, 2],
     xlab = "Soft Threshold (power)",
     ylab = "Scale Free Topology Model Fit, signed R^2",
     type = "n", main = "Scale independence")
text(sft$fitIndices[, 1],
     -sign(sft$fitIndices[, 3]) * sft$fitIndices[, 2],
     labels = powers, cex = cex1, col = "red")
abline(h = 0.70, col = "red")

# Mean connectivity plot
plot(sft$fitIndices[, 1], sft$fitIndices[, 5],
     xlab = "Soft Threshold (power)",
     ylab = "Mean Connectivity",
     type = "n", main = "Mean connectivity")
text(sft$fitIndices[, 1], sft$fitIndices[, 5],
     labels = powers, cex = cex1, col = "red")
```

## Network construction and module detection

Construct a signed co-expression network using `blockwiseModules` and detect
modules of co-occurring ASVs.

```{r construct-network}
# Set soft threshold power
# NOTE: softPower=12 is carried forward from the original analysis.
# Verify that R^2 >= 0.70 at this power before proceeding.
softPower <- 12

# Check R^2 at chosen power
sft_row <- which(sft$fitIndices[, 1] == softPower)
if (length(sft_row) == 1) {
  sft_r2 <- -sign(sft$fitIndices[sft_row, 3]) * sft$fitIndices[sft_row, 2]
  message(sprintf("Scale-free fit R^2 at power %d: %.3f", softPower, sft_r2))
  if (sft_r2 < 0.70) {
    warning(sprintf(
      "R^2 = %.3f at power %d is below the 0.70 threshold. Consider a different power.",
      sft_r2, softPower))
  }
}

# Construct network and detect modules
net <- blockwiseModules(tcounts,
                        power = softPower,
                        networkType = "signed",
                        TOMType = "signed",
                        minModuleSize = 30,
                        reassignThreshold = 0,
                        mergeCutHeight = 0.25,
                        numericLabels = TRUE,
                        pamRespectsDendro = FALSE,
                        verbose = 3)

moduleColors <- labels2colors(net$colors)
MEs_raw <- net$MEs  # preserve original MEs before reordering

# Print module sizes for reference
message("Module sizes (after merging):")
print(table(moduleColors))
```

## Module eigengene dendrogram

```{r me-dendrogram, fig.width=10, fig.height=6}
## Figure S11
moduleTree <- hclust(dist(1 - cor(MEs_raw)), method = "average")

# Map ME labels to WGCNA colors for the dendrogram
me_numeric_labels <- as.numeric(sub("^ME", "", colnames(MEs_raw)))
me_colors <- labels2colors(me_numeric_labels)
moduleTree$labels <- paste0(colnames(MEs_raw), " (", me_colors, ")")

plotDendroAndColors(moduleTree,
                    colors = me_colors,
                    groupLabels = "Module color",
                    dendroLabels = moduleTree$labels,
                    main = "Clustering of module eigengenes",
                    cex.dendroLabels = 1.0,
                    cex.colorLabels = 0.8)
```

# Module–Trait Correlation

Correlate module eigengenes with all measured environmental traits (nutrients,
temperature, pH, dissolved oxygen, microcystin).

```{r module-trait-heatmap, fig.width=10, fig.height=10}
## Figure 8
nTaxa <- ncol(tcounts)
nSamples <- nrow(tcounts)

# Recalculate ordered module eigengenes for the heatmap
MEs0 <- moduleEigengenes(tcounts, moduleColors)$eigengenes
MEs <- orderMEs(MEs0)

# Compute correlations and p-values between MEs and all numeric traits
moduleTraitCor <- cor(MEs, datTraits, use = "p")
moduleTraitPvalue <- corPvalueStudent(moduleTraitCor, nSamples)

# Plot labeled heatmap
textMatrix <- paste(signif(moduleTraitCor, 2), "\n(",
                    signif(moduleTraitPvalue, 1), ")", sep = "")
dim(textMatrix) <- dim(moduleTraitCor)

par(mar = c(6, 8.5, 3, 3))
labeledHeatmap(Matrix = moduleTraitCor,
               xLabels = names(datTraits),
               yLabels = names(MEs),
               ySymbols = names(MEs),
               colorLabels = FALSE,
               colors = blueWhiteRed(50),
               textMatrix = textMatrix,
               setStdMargins = FALSE,
               cex.text = 0.8,
               zlim = c(-1, 1),
               main = "Module-trait relationships")
```

# Intramodular Connectivity and Hub Taxa

Identify hub ASVs (highly connected nodes within each module).

```{r hub-taxa}
# Calculate adjacency matrix (same parameters as blockwiseModules)
adjacency <- adjacency(tcounts, power = softPower, type = "signed")

# Calculate intramodular connectivity
kIN <- intramodularConnectivity(adjacency, moduleColors)

# Combine connectivity with ASV and module information
ASV_Info <- data.frame(
  ASV = names(tcounts),
  Module_Color = moduleColors,
  kIN_Within = kIN$kWithin
)

# Identify hub taxa in key modules
# NOTE: Replace these with the modules of interest from section 4 results
key_modules <- c("blue", "pink")

hub_taxa_list <- ASV_Info %>%
  filter(Module_Color %in% key_modules) %>%
  group_by(Module_Color) %>%
  arrange(desc(kIN_Within))

top_10_hubs <- hub_taxa_list %>%
  slice_head(n = 10) %>%
  as.data.frame()

print("Top 10 Hub Taxa (ASVs) in Key Modules:")
print(top_10_hubs)

write.csv(ASV_Info, "results/wgcna/ASV_Intramodular_Connectivity.csv",
          row.names = FALSE)
```

# Taxonomy Lookup and Module Membership

## Build taxonomy lookup table

```{r taxonomy-lookup}
# Helper: extract taxonomy columns from a Table SX file
extract_taxa_from_table_sx <- function(file_path, domain_tag, taxa_cols) {
  raw <- read.csv(file = file_path, header = TRUE, row.names = 1,
                  check.names = FALSE)
  taxa_df <- raw[, intersect(colnames(raw), taxa_cols), drop = FALSE]
  taxa_df$ASV_ID <- paste0(domain_tag, "_", rownames(taxa_df))
  return(taxa_df)
}

taxa_16s_raw <- extract_taxa_from_table_sx(
  "data/Table_SX_16S_2023_ASV_Count_and_Taxonomy.csv", "16S", taxa_cols_16s)
taxa_18s_raw <- extract_taxa_from_table_sx(
  "data/Table_SX_18S_2023_ASV_Count_and_Taxonomy.csv", "18S", taxa_cols_18s)

# Standardize: rename "Taxa" -> "Taxa2", keep shared columns
standardize_taxa <- function(df) {
  colnames(df)[colnames(df) == "Taxa"] <- "Taxa2"
  keep_cols <- intersect(colnames(df),
                         c("ASV_ID", "Class", "Order", "Family",
                           "Genus", "Species", "Taxa2"))
  return(df[, keep_cols, drop = FALSE])
}

taxa_16s_std <- standardize_taxa(taxa_16s_raw)
taxa_18s_std <- standardize_taxa(taxa_18s_raw)

ASV_taxa <- rbind(taxa_16s_std, taxa_18s_std)
rownames(ASV_taxa) <- ASV_taxa$ASV_ID
```

## Module Membership (MM)

Module membership quantifies how strongly each ASV correlates with its module
eigengene.

```{r module-membership}
modNames <- substring(names(MEs), 3)

TaxaModuleMembership <- as.data.frame(cor(tcounts, MEs, use = "p"))
MMPvalue <- as.data.frame(corPvalueStudent(as.matrix(TaxaModuleMembership),
                                           nSamples))
names(TaxaModuleMembership) <- paste0("MM", modNames)
names(MMPvalue) <- paste0("p.MM", modNames)
```

# Module–Trait Exploration

## Define exploration function

The `explore_module_trait()` function computes Gene Significance (GS) for a
given trait, pairs it with Module Membership (MM), annotates with taxonomy, and
produces a scatter plot.

```{r define-explore-function}
explore_module_trait <- function(module_color,
                                 trait_column_name,
                                 datTraits_data,
                                 tcounts_data,
                                 moduleColors_data,
                                 ASV_taxa_data,
                                 alpha = 0.05,
                                 highlight_taxa = NULL,
                                 highlight_name = NULL) {
  
  # Compute nSamples locally
  n_samples <- nrow(tcounts_data)
  
  # 1. Prepare trait vector
  trait_data <- as.data.frame(datTraits_data[[trait_column_name]])
  names(trait_data) <- trait_column_name
  
  # 2. Calculate Gene Significance (GS) for this trait
  TaxaTraitSignificance <- as.data.frame(cor(tcounts_data, trait_data, use = "p"))
  GSPvalue <- as.data.frame(corPvalueStudent(as.matrix(TaxaTraitSignificance),
                                             n_samples))
  names(TaxaTraitSignificance) <- paste0("GS.", trait_column_name)
  names(GSPvalue) <- paste0("p.GS.", trait_column_name)
  
  # 3. Find the MM column for this module
  column <- match(module_color, modNames)
  if (is.na(column)) stop(paste("Module color", module_color, "not found in modNames."))
  
  # 4. Assemble data frame with MM, GS, and taxonomy
  taxa_module_trait_data <- data.frame(
    ASV_ID = rownames(TaxaModuleMembership),
    Module_Color = moduleColors_data,
    MM = TaxaModuleMembership[, column],
    p.MM = MMPvalue[, column],
    GS = TaxaTraitSignificance[, 1],
    p.GS = GSPvalue[, 1],
    stringsAsFactors = FALSE
  )
  
  final_taxa_table <- merge(taxa_module_trait_data, ASV_taxa_data,
                            by = "ASV_ID", all.x = TRUE)
  
  # Subset to the target module
  module_output <- final_taxa_table[final_taxa_table$Module_Color == module_color, ]
  
  # Sort by |GS| then |MM| descending
  module_output <- module_output[order(abs(module_output$GS),
                                       abs(module_output$MM),
                                       decreasing = TRUE), ]
  
  # Add trait name column for downstream merging
  module_output$Trait <- trait_column_name
  
  # 5. Determine point colors for plot
  is_sig <- module_output$p.MM < alpha & module_output$p.GS < alpha
  point_colors <- rep("gray70", nrow(module_output))
  
  if (!is.null(highlight_taxa)) {
    # Highlight a specific taxon
    is_highlight <- is_sig & (module_output$Taxa2 == highlight_taxa)
    point_colors[is_highlight] <- module_color
    
    legend_data <- list(
      labels = c(paste(highlight_name %||% highlight_taxa, "(p <", alpha, ")"),
                 "Other Taxa"),
      colors = c(module_color, "gray70")
    )
    title_text <- paste(highlight_name %||% highlight_taxa, "significance")
    
  } else {
    # Color all significant points by Taxa2
    taxa_vec <- module_output$Taxa2
    sig_categories <- unique(taxa_vec[is_sig])
    n_cat <- length(sig_categories)
    
    if (n_cat > 0) {
      get_colors <- colorRampPalette(brewer.pal(max(n_cat, 3), "Set1"))
      taxa_colors <- setNames(get_colors(n_cat), sig_categories)
      point_colors[is_sig] <- taxa_colors[taxa_vec[is_sig]]
    } else {
      taxa_colors <- c()
    }
    
    legend_data <- list(labels = names(taxa_colors), colors = taxa_colors)
    title_text <- "Taxa membership by taxonomic group"
  }
  
  # 6. Export results
  output_filename <- paste0("Taxa_Data_", module_color,
                            "_Module_vs_", trait_column_name, "_with_Taxa.csv")
  write.csv(module_output,
            file = file.path("results/wgcna", output_filename),
            row.names = FALSE)
  
  # 7. Plot MM vs GS scatter
  par(mfrow = c(1, 1))
  verboseScatterplot(
    abs(module_output$MM),
    abs(module_output$GS),
    xlab = paste("Module Membership in", module_color, "module (|MM|)"),
    ylab = paste("Taxa significance for", trait_column_name, "(|GS|)"),
    main = paste("MM vs. GS in", module_color, "module:", title_text),
    cex.main = 1.2, cex.lab = 1.2, cex.axis = 1.2,
    col = point_colors,
    pch = 16,
    cex = ifelse(is_sig, 1.2, 0.8)
  )
  
  if (length(legend_data$labels) > 0 && !is.null(legend_data$labels[1])) {
    legend("topleft",
           legend = legend_data$labels,
           col = legend_data$colors,
           pch = 16, bty = "n", cex = 0.8)
  }
}
```

## Apply to specific module–trait combinations

> **NOTE:** Only 2 module–trait combinations are shown here as examples. Add
> additional calls to `explore_module_trait()` as needed for your full analysis.

```{r explore-black-bdo, fig.width=8, fig.height=8}
explore_module_trait(
  module_color     = "black",
  trait_column_name = "Bdo",
  datTraits_data   = datTraits,
  tcounts_data     = tcounts,
  moduleColors_data = moduleColors,
  ASV_taxa_data    = ASV_taxa
)
```

```{r explore-brown-stemp, fig.width=8, fig.height=8}
explore_module_trait(
  module_color     = "brown",
  trait_column_name = "stemp",
  datTraits_data   = datTraits,
  tcounts_data     = tcounts,
  moduleColors_data = moduleColors,
  ASV_taxa_data    = ASV_taxa
)
```

# Merge Module–Trait Results

Combine all per-module-trait CSV outputs into a single summary table.

```{r merge-module-trait-results}
output_dir <- "results/wgcna/"
file_pattern <- "^Taxa_Data_.*_with_Taxa\\.csv$"

all_files <- list.files(path = output_dir, pattern = file_pattern,
                        full.names = TRUE)

if (length(all_files) == 0) {
  stop("No module-trait CSV files found in 'results/wgcna/'. Ensure section 8 ran successfully.")
}

# Use bind_rows to handle potential column mismatches from stale files
data_list <- lapply(all_files, read.csv, stringsAsFactors = FALSE)
combined_trait_data <- bind_rows(data_list)

write.csv(combined_trait_data,
          file = paste0(output_dir, "Combined_All_Module_Trait_Results.csv"),
          row.names = FALSE)

message(sprintf("Merged %d files into Combined_All_Module_Trait_Results.csv (%d rows).",
                length(all_files), nrow(combined_trait_data)))
```

# Export Module Assignments

```{r export-module-assignments}
# Module assignments
asv_module_table <- data.frame(
  ASV = colnames(tcounts),
  Module_Color = moduleColors,
  stringsAsFactors = FALSE
)
asv_module_table <- asv_module_table[order(asv_module_table$Module_Color), ]
write.csv(asv_module_table, file = "results/wgcna/asv_module_assignments.csv",
          row.names = FALSE)

# Module assignments merged with sample-level Hellinger-transformed counts
tcounts_transposed <- as.data.frame(t(tcounts))
asv_module_samples <- merge(asv_module_table, tcounts_transposed,
                            by.x = "ASV", by.y = "row.names")
asv_module_samples <- asv_module_samples[, c("ASV", "Module_Color",
                                             setdiff(names(asv_module_samples),
                                                     c("ASV", "Module_Color")))]
write.csv(asv_module_samples,
          file = "results/wgcna/asv_module_and_sample_data.csv",
          row.names = FALSE)
```

# Inter-Domain (16S vs 18S) Correlation Analysis

Calculate cross-domain correlations within each WGCNA module to identify
significant prokaryote–eukaryote associations.

```{r interdomain-correlations}
# --- Determine ASV domain membership ---
get_asv_type <- function(asv_id, counts_16s_df, counts_18s_df) {
  if (asv_id %in% rownames(counts_16s_df)) return("16S")
  if (asv_id %in% rownames(counts_18s_df)) return("18S")
  return("Unknown")
}

asv_ids_in_wgcna <- colnames(tcounts)
asv_domains <- sapply(asv_ids_in_wgcna, function(id) {
  get_asv_type(id, counts_filt16s, counts_filt18s)
})
asv_domain_df <- data.frame(ASV_ID = asv_ids_in_wgcna,
                            Domain = asv_domains,
                            row.names = NULL,
                            stringsAsFactors = FALSE)

# --- Inter-domain correlation function ---
run_all_interdomain_correlations <- function(tcounts_data,
                                             net_colors_data,
                                             asv_domain_data,
                                             TaxaModuleMembership,
                                             MMPvalue,
                                             ASV_taxa_data,
                                             alpha = 0.05) {
  
  unique_module_labels <- unique(net_colors_data)
  unique_module_labels <- unique_module_labels[unique_module_labels != 0]
  all_sig_cors <- list()
  nSamples_local <- nrow(tcounts_data)
  
  all_color_map <- labels2colors(unique_module_labels)
  names(all_color_map) <- unique_module_labels
  
  for (module_label in unique_module_labels) {
    
    module_color <- all_color_map[as.character(module_label)]
    if (is.null(module_color) || is.na(module_color)) next
    
    message(sprintf("Processing module %s (color: %s)", module_label, module_color))
    
    # Isolate ASVs in this module
    module_ASVs <- names(net_colors_data[net_colors_data == module_label])
    
    # Split by domain
    module_domains <- asv_domain_data[asv_domain_data$ASV_ID %in% module_ASVs, ]
    asv_16s <- module_domains$ASV_ID[module_domains$Domain == "16S"]
    asv_18s <- module_domains$ASV_ID[module_domains$Domain == "18S"]
    
    if (length(asv_16s) == 0 || length(asv_18s) == 0) {
      message(sprintf("  Skipping %s: 16S=%d, 18S=%d", module_color,
                      length(asv_16s), length(asv_18s)))
      next
    }
    
    # Calculate cross-domain correlation matrix
    data_16s <- tcounts_data[, asv_16s, drop = FALSE]
    data_18s <- tcounts_data[, asv_18s, drop = FALSE]
    
    cor_matrix <- cor(data_16s, data_18s, use = "pairwise.complete.obs")
    pval_matrix <- corPvalueStudent(cor_matrix, nSamples_local)
    
    # Convert to long format
    cor_df <- data.frame(
      ASV_16S = rep(rownames(cor_matrix), times = ncol(cor_matrix)),
      ASV_18S = rep(colnames(cor_matrix), each = nrow(cor_matrix)),
      Module_Color = module_color,
      Correlation_16S_18S = as.vector(cor_matrix),
      P_value_16S_18S = as.vector(pval_matrix),
      stringsAsFactors = FALSE
    )
    
    # Filter to significant correlations
    sig_df <- cor_df[cor_df$P_value_16S_18S < alpha, ]
    message(sprintf("  Found %d significant pairs (p < %g)", nrow(sig_df), alpha))
    
    if (nrow(sig_df) > 0) {
      sig_df$ASV_16S <- as.character(sig_df$ASV_16S)
      sig_df$ASV_18S <- as.character(sig_df$ASV_18S)
      
      # Add module membership values
      mm_col <- paste0("MM", module_label)
      pmm_col <- paste0("p.MM", module_label)
      
      sig_df$MM_16S   <- TaxaModuleMembership[match(sig_df$ASV_16S, rownames(TaxaModuleMembership)), mm_col]
      sig_df$p.MM_16S <- MMPvalue[match(sig_df$ASV_16S, rownames(MMPvalue)), pmm_col]
      sig_df$MM_18S   <- TaxaModuleMembership[match(sig_df$ASV_18S, rownames(TaxaModuleMembership)), mm_col]
      sig_df$p.MM_18S <- MMPvalue[match(sig_df$ASV_18S, rownames(MMPvalue)), pmm_col]
      
      # Add taxonomy info
      taxa_16s <- ASV_taxa_data[match(sig_df$ASV_16S, ASV_taxa_data$ASV_ID),
                                c("Taxa2", "ASV_ID")]
      taxa_18s <- ASV_taxa_data[match(sig_df$ASV_18S, ASV_taxa_data$ASV_ID),
                                c("Taxa2", "ASV_ID")]
      colnames(taxa_16s) <- paste0("16S_", colnames(taxa_16s))
      colnames(taxa_18s) <- paste0("18S_", colnames(taxa_18s))
      
      sig_df <- cbind(sig_df, taxa_16s, taxa_18s)
      all_sig_cors[[module_color]] <- sig_df
    }
  }
  
  # Combine all modules
  final_table <- do.call(rbind, all_sig_cors)
  if (is.null(final_table)) {
    message("No significant inter-domain correlations found.")
    return(NULL)
  }
  
  # Sort by module then by correlation strength
  final_table <- final_table[order(final_table$Module_Color,
                                   -abs(final_table$Correlation_16S_18S)), ]
  
  write.csv(final_table,
            file = "results/wgcna/All_InterDomain_Correlations_Combined.csv",
            row.names = FALSE)
  message(sprintf("Inter-domain analysis complete: %d significant pairs saved.",
                  nrow(final_table)))
  
  return(final_table)
}

# Run the inter-domain analysis
all_interdomain_cors <- run_all_interdomain_correlations(
  tcounts_data = tcounts,
  net_colors_data = net$colors,
  asv_domain_data = asv_domain_df,
  TaxaModuleMembership = TaxaModuleMembership,
  MMPvalue = MMPvalue,
  ASV_taxa_data = ASV_taxa,
  alpha = 0.05
)
```

## Filter inter-domain correlations (stricter thresholds)

Apply more stringent cutoffs (p < 0.01, |r| > 0.4) for downstream network
visualization.

```{r filter-interdomain}
input_file  <- "results/wgcna/All_InterDomain_Correlations_Combined.csv"
output_file <- "results/wgcna/Filtered_InterDomain_Correlations_R.csv"

data_df <- read_csv(input_file, show_col_types = FALSE)

# Apply stricter thresholds
filtered_df <- data_df %>%
  filter(P_value_16S_18S < 0.01,
         abs(Correlation_16S_18S) > 0.4)

message(sprintf("Inter-domain filtering: %d -> %d correlations (p<0.01, |r|>0.4)",
                nrow(data_df), nrow(filtered_df)))

write_csv(filtered_df, output_file)
```

# Module–Trait–Taxa Summary and Bubble Chart

Identify the most abundant taxon per module–trait combination and visualize
using a bubble chart.

```{r bubble-chart, fig.width=10, fig.height=8}
## Figure 9
original_data <- read_csv("results/wgcna/Combined_All_Module_Trait_Results.csv",
                          show_col_types = FALSE)

# Find the most frequent Taxa2 per module-trait combination
# and compute its average Gene Significance
top_taxa_gs_summary <- original_data %>%
  group_by(Module_Color, Trait, Taxa2) %>%
  summarise(Count = n(), Avg_GS = mean(GS), .groups = "drop_last") %>%
  mutate(Total_Count = sum(Count)) %>%
  arrange(desc(Count)) %>%
  slice_head(n = 1) %>%
  ungroup() %>%
  mutate(Proportion = Count / Total_Count) %>%
  select(Module_Color, Trait, Top_Taxa2 = Taxa2,
         Count, Total_Count, Proportion, Avg_GS)

write_csv(top_taxa_gs_summary, "Top_Taxa_Per_Module_Trait_with_GS.csv")

# Prepare plot data
data_for_plot <- top_taxa_gs_summary %>%
  mutate(
    Correlation_Direction = case_when(
      Avg_GS > 0  ~ "Positive",
      Avg_GS < 0  ~ "Negative",
      TRUE         ~ "Neutral"
    ),
    Correlation_Strength = abs(Avg_GS)
  )

# Bubble chart
p_enhanced <- data_for_plot %>%
  mutate(Top_Taxa2 = factor(Top_Taxa2)) %>%
  ggplot(aes(x = Trait, y = Module_Color)) +
  geom_point(aes(fill = Top_Taxa2,
                 size = Correlation_Strength,
                 color = Correlation_Direction),
             shape = 21, stroke = 1.5, alpha = 0.8) +
  scale_size_continuous(name = "|Average GS| (Strength)", range = c(2, 15)) +
  scale_color_manual(name = "Avg. GS Direction",
                     values = c("Positive" = "darkred",
                                "Negative" = "darkblue",
                                "Neutral" = "gray")) +
  labs(title = "Top Taxa within Each Module-Trait Combination",
       subtitle = "Size = |Avg. GS|, Border = GS direction, Fill = Top taxon",
       x = "Trait", y = "Module Color") +
  theme_bw(base_size = 16) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 14),
        axis.text.y = element_text(size = 14),
        axis.title = element_text(size = 16),
        plot.title = element_text(hjust = 0.5, face = "bold", size = 18),
        plot.subtitle = element_text(hjust = 0.5, size = 14),
        legend.text = element_text(size = 12),
        legend.title = element_text(size = 13),
        panel.grid.minor = element_blank(),
        legend.box = "vertical")

print(p_enhanced)
```

# Inter-Domain Network Visualization

Visualize the filtered inter-domain correlations as a network graph, with
separate plots per WGCNA module.

```{r network-plots, fig.width=8, fig.height=8, results='asis'}
# Load filtered inter-domain correlations
df <- read_csv("results/wgcna/Filtered_InterDomain_Correlations_R.csv",
               show_col_types = FALSE)

# Define stricter thresholds for network visualization
P_cutoff <- 0.001
R_cutoff <- 0.7

# Filter to keep only strong and significant edges
full_filtered_df <- df %>%
  filter(
    P_value_16S_18S < P_cutoff,
    abs(Correlation_16S_18S) >= R_cutoff
  ) %>%
  select(ASV_18S, ASV_16S, Module_Color, Correlation_16S_18S,
         P_value_16S_18S, `18S_Taxa2`, `16S_Taxa2`) %>%
  rename(
    from = ASV_18S,
    to = ASV_16S,
    weight = Correlation_16S_18S
  )

cat(sprintf("Total edges after filtering: %d\n\n", nrow(full_filtered_df)))

# Build full node list
nodes_18S <- full_filtered_df %>% 
  select(name = from, Taxa = `18S_Taxa2`, Module_Color) %>% 
  mutate(Domain = "Eukaryote (18S)", Label = name)

nodes_16S <- full_filtered_df %>% 
  select(name = to, Taxa = `16S_Taxa2`, Module_Color) %>% 
  mutate(Domain = "Prokaryote (16S)", Label = name)

full_nodes <- bind_rows(nodes_18S, nodes_16S) %>%
  distinct(name, .keep_all = TRUE) 

module_list <- unique(full_nodes$Module_Color)

# Loop through each module and plot separately
for (current_module in module_list) {
  
  cat(sprintf("\n\n## Module: %s\n\n", current_module))
  
  # Filter data for the current module
  module_nodes <- full_nodes %>% 
    filter(Module_Color == current_module)
  
  module_edges <- full_filtered_df %>%
    filter(from %in% module_nodes$name & to %in% module_nodes$name)
  
  if (nrow(module_edges) == 0) {
    cat(sprintf("Module %s has no internal connections above threshold. Skipping.\n\n",
                current_module))
    next
  }
  
  # Create igraph object
  module_graph <- graph_from_data_frame(
    d = module_edges, 
    vertices = module_nodes, 
    directed = FALSE
  )
  
  # Visualize the module network
  network_plot <- ggraph(module_graph, layout = 'fr') + 
    geom_edge_fan(
      aes(color = weight > 0, width = abs(weight), alpha = abs(weight)),
      show.legend = TRUE
    ) +
    scale_edge_width_continuous(range = c(0.2, 1), name = "|Correlation|") +
    scale_edge_color_manual(
      values = c("FALSE" = "blue", "TRUE" = "red"),
      labels = c("Negative", "Positive"),
      name = "Correlation Sign"
    ) +
    geom_node_point(
      aes(color = Taxa, shape = Domain, size = 2)
    ) +
    scale_shape_manual(
      values = c("Eukaryote (18S)" = 17, "Prokaryote (16S)" = 19),
      name = "Domain"
    ) +
    geom_node_text(aes(label = Label), repel = TRUE, size = 2) +
    labs(
      title = paste0("Module: ", current_module,
                     " Network (N=", vcount(module_graph), " ASVs)"),
      subtitle = paste0("Filtered: |r| > ", R_cutoff, " and P < ", P_cutoff)
    ) +
    theme_graph()
  
  print(network_plot)
  
  # Save SVG
  filename <- paste0("figs/Network_Module_", current_module, ".svg")
  ggsave(filename, plot = network_plot, width = 8, height = 8, dpi = 600)
}
```

# Core Taxa–Environment Time Series

Plot the temporal dynamics of Planktothrix and a correlated environmental
variable (sdo) across lakes, for the yellow and blue WGCNA modules.

```{r define-taxa-processing}
# Helper function to process taxa data for time series plotting
process_taxa_data <- function(file_path) {
  read_csv(file_path, show_col_types = FALSE) %>%
    filter(Taxa %in% c("Planktothrix", "sdo")) %>%
    pivot_longer(
      cols = starts_with(c("LM", "LR", "TTL")),
      names_to = "Sample_ID",
      values_to = "Counts"
    ) %>%
    mutate(
      Lake = str_extract(Sample_ID, "^(LM|LR|TTL)"),
      Date_String = str_extract(Sample_ID, "\\d{1}-\\d{1,2}$"),
      Date = as.Date(paste0("2024-", Date_String), format = "%Y-%m-%d")
    ) %>%
    group_by(Taxa, Lake, Date) %>%
    summarise(
      Mean_Count = mean(Counts, na.rm = TRUE),
      SD_Count = sd(Counts, na.rm = TRUE),
      .groups = 'drop'
    )
}
```

```{r core-taxa-plots, fig.width=6.5, fig.height=9}
# Load data for yellow and blue modules
yellow_summary <- process_taxa_data("data/correlations_yellow_sdo.csv")
blue_summary   <- process_taxa_data("data/correlations_blue_sdo.csv")

# --- Plot A: Yellow Module ---
plot_yellow <- yellow_summary %>%
  ggplot(aes(x = Date, y = Mean_Count, color = Taxa, group = Taxa)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  geom_errorbar(aes(ymin = Mean_Count - SD_Count,
                    ymax = Mean_Count + SD_Count),
                width = 0, linewidth = 0.5) +
  facet_grid(Taxa ~ Lake, scales = "free_y") +
  scale_color_manual(values = c("Planktothrix" = rgb(218, 185, 46,
                                                     maxColorValue = 255),
                                "NP" = "grey50")) +
  scale_x_date(date_labels = "%m-%d", date_breaks = "2 weeks") +
  labs(title = "Yellow Module",
       x = "Sampling Date",
       y = "Average ASV Count (Mean +/- SD)",
       color = "Taxa") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "none")

# --- Plot B: Blue Module ---
plot_blue <- blue_summary %>%
  ggplot(aes(x = Date, y = Mean_Count, color = Taxa, group = Taxa)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  geom_errorbar(aes(ymin = Mean_Count - SD_Count,
                    ymax = Mean_Count + SD_Count),
                width = 0, linewidth = 0.5) +
  facet_grid(Taxa ~ Lake, scales = "free_y") +
  scale_color_manual(values = c("Planktothrix" = rgb(218, 185, 46,
                                                     maxColorValue = 255),
                                "NP" = "grey50")) +
  scale_x_date(date_labels = "%m-%d", date_breaks = "2 weeks") +
  labs(title = "Blue Module",
       x = "Sampling Date",
       y = "Average ASV Count (Mean +/- SD)",
       color = "Taxa") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "none")

# --- Combine and label A/B ---
combined_plot <- plot_yellow / plot_blue + 
  plot_annotation(tag_levels = 'A')

print(combined_plot)

ggsave("figs/combined_modules_sdo.png", combined_plot,
       width = 6.5, height = 9, units = "in")
```

# Session Info

```{r session-info}
sessionInfo()
```
